### 2024-11-18 paper/paper.qmd
Initial Commit - paper and readme added

---
title: "Himalays"
# subtitle: "My subtitle if needed"
author: 
  - Khushaal Nandwani
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
# bibliography: references.bib
---

# Possible topics

1. Predicting the average height the person will be able to climb, or the percentage of heigh depending on the various variables.

# Lemur Data

1. What factors can contribute to lifespan of lemur?
    - sex
    - species
    - birth month (season): some seasons makes stronger. maybe winter?
    - captivity vs wild: captivity makes them use to spoon feeding?
    - litter size: maybe mother is more involved with other kids
    - mother_species and father_species is same or not

## Why Lemur?

Established in 1966 on Duke University's campus in Durham, NC, the Duke Lemur Center (DLC) is a global leader in the research, care, and conservation of lemurs, the planet's most endangered group of mammals. Home to over 200 animals spanning 13 species, the DLC hosts the most diverse population of lemurs outside their native habitat in Madagascar. cite(https://lemur.duke.edu)

### 2024-11-18 scripts/00-install_packages.r
clean scripts and simulation scripts added

@@ -1,0 +1,22 @@
+#### Preamble ####
+# Purpose: Install the required packages for the analysis
+# Author: Khushaal Nandwani
+# Date: 18 November 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+
+# List of packages to install
+packages <- c(
+  "tidyverse",
+  "dplyr",
+  "knitr",
+  "ggplot2",
+)
+
+# Install packages if they are not already installed
+for (package in packages) {
+  if (!requireNamespace(package, quietly = TRUE)) {
+    install.packages(package, dependencies = TRUE)
+  }
+}

### 2024-11-18 scripts/01-simulate_data.r

@@ -1,0 +1,66 @@
+#### Preamble ####
+# Purpose: Simulated the analysis dataset
+# Author: Khushaal Nandwani
+# Date: 18 November 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+
+## Workspace Setup ##
+library(tidyverse)
+
+## Simulation ##
+# set seed for reproducibility
+set.seed(136)
+
+# number of rows in the dataset
+n <- 1200
+
+# generate animal ids
+animal_id <- 1:n
+
+# generate species
+types_of_species <- c(
+    "Northern greater galago",
+    "Collared brown lemur",
+    "Eulemur hybrid",
+    "Red-fronted brown lemur",
+    "Mohol bushbaby",
+    "Eastern lesser bamboo lemur",
+    "Black lemur",
+    "Common brown lemur"
+)
+species <- sample(types_of_species, n, replace = TRUE)
+
+# generate sex
+sex <- sample(c("M", "F"), n, replace = TRUE)
+
+# generate birth_type
+birth_type <- sample(c("captive", "wild"), n, replace = TRUE)
+
+# generate litter_size
+litter_size <- sample(1:5, n, replace = TRUE)
+
+# generate mother_father_same
+mother_father_same <- sample(c(0, 1), n, replace = TRUE)
+
+# generate age, making it dependent on birth_type
+age <- ifelse(birth_type == "captive", rnorm(n, 9, 2), rnorm(n, 7, 2))
+
+# generate month_born
+month_born <- sample(1:12, n, replace = TRUE)
+
+# create the dataset
+data <- tibble(
+    animal_id = animal_id,
+    species = species,
+    sex = sex,
+    birth_type = birth_type,
+    litter_size = litter_size,
+    mother_father_same = mother_father_same,
+    age = age,
+    month_born = month_born
+)
+
+# print the first few rows of the dataset
+head(data)

### 2024-11-18 scripts/02-clean_data.r

@@ -1,0 +1,45 @@
+#### Preamble ####
+# Purpose: Cleans the data and keeps only the relevant columns
+# Author: Khushaal Nandwani
+# Date: 18 November 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The raw data file should be present in the data/raw_data folder
+
+# Most of the data from raw data was already cleaned by Cook
+# However, to conduct our analysis, we will only keep the columns we need
+
+## Workspace Setup ##
+library(tidyverse)
+
+w <- read.csv("finalPaper/data/raw_data/weights.csv")
+
+## Data Cleaning ##
+# read the data
+data <- read.csv("finalPaper/data/raw_data/animals.csv")
+
+# selecting relevant columns
+data <- data[, c("animal_id", "species", "sex", "birth_date", "death_date", "birth_type", "litter_size", "mother_species", "father_species")]
+
+# create a column, which says if mother and father species is same
+data$mother_father_same <- ifelse(data$mother_species == data$father_species, 1, 0)
+
+# remove the mother_species and father_species columns, as they are no longer needed
+data <- data[, -c(8, 9)]
+
+# create a column for age
+data$birth_date <- as.Date(data$birth_date)
+data$death_date <- as.Date(data$death_date)
+data$age <- as.numeric(difftime(data$death_date, data$birth_date, units = "days"))/365
+
+# create a column for month born
+data$month_born <- as.numeric(format(data$birth_date, "%m"))
+
+# remove birth_date and death_date columns, as they are no longer needed
+data <- data[, -c(4, 5)]
+
+# remove rows with missing values except missing in litter_size and mother_father_same
+data <- data %>% drop_na(animal_id, species, sex, birth_type, age, month_born)
+
+# save the cleaned data
+write.csv(data, "finalPaper/data/analysis_data/analysis_data.csv", row.names = FALSE)

### 2024-11-23 scripts/01-simulate_data.r
scripts and data for wild and captive added

@@ -62,5 +62,7 @@ data <- tibble(
     month_born = month_born
 )
 
+write.csv(data, "data/analysis_data/sim_data.csv", row.names = FALSE)
+
 # print the first few rows of the dataset
 head(data)

### 2024-11-23 scripts/02-clean_data.r

@@ -4,22 +4,32 @@
 # Date: 18 November 2024
 # Contact: khushaal.nandwani@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The raw data file should be present in the data/raw_data folder
-
-# Most of the data from raw data was already cleaned by Cook
-# However, to conduct our analysis, we will only keep the columns we need
+# Pre-requisites: The raw data file should be present in the data/raw_data folde
 
 ## Workspace Setup ##
 library(tidyverse)
 
-w <- read.csv("finalPaper/data/raw_data/weights.csv")
+## Load the data ##
+# REMOVE: w <- read.csv("data/raw_data/weights.csv")
 
 ## Data Cleaning ##
-# read the data
-data <- read.csv("finalPaper/data/raw_data/animals.csv")
+data <- read.csv("data/raw_data/animals.csv")
+
+# get unique values of taxonomic code and number of unique values
+unique(data$taxonomic_code)
+# species
+unique(data$species)
 
 # selecting relevant columns
-data <- data[, c("animal_id", "species", "sex", "birth_date", "death_date", "birth_type", "litter_size", "mother_species", "father_species")]
+data <- data[ , c("animal_id", "taxonomic_code", "sex", "birth_date", "death_date", "birth_type", "litter_size", "mother_species", "father_species")]
+
+# make first letter of taxnomic code as Genus
+# the rest of the letters as species
+
+data$species <- substr(data$taxonomic_code, 1, 1)
+data$genus <- substr(data$taxonomic_code, 2, nchar(data$taxonomic_code))
+
+
 
 # create a column, which says if mother and father species is same
 data$mother_father_same <- ifelse(data$mother_species == data$father_species, 1, 0)
@@ -37,9 +47,24 @@ data$month_born <- as.numeric(format(data$birth_date, "%m"))
 
 # remove birth_date and death_date columns, as they are no longer needed
 data <- data[, -c(4, 5)]
+# splilt the data into birth_type == captive or wild
+data_captive <- data[data$birth_type == "captive", ]
 
+data_wild <- data[data$birth_type == "wild", ]
+# since there is no data on the mothr father and litter size of iwld data, we remove these cols
+data_wild <- data_wild[, -c(5, 6)]
+head(data_wild)
 # remove rows with missing values except missing in litter_size and mother_father_same
-data <- data %>% drop_na(animal_id, species, sex, birth_type, age, month_born)
[…]

### 2024-11-23 scripts/04-data-analysis.r

@@ -1,0 +1,41 @@
+## Preamble ##
+# Purpose: Data analysis for the project. Specifically, we will build a model to predict the age of animals based on various factors.
+# Author: Khushaal Nandwani
+# Date: 18 November 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+
+count_na <- function(data) {
+    sapply(data, function(x) sum(is.na(x)))
+}
+## Workspace setup ##
+library(tidyverse)
+
+## Modeling ##
+# Read the cleaned analysis dataset
+# analysis_data <- read.csv("data/analysis_data/analysis_data.csv")
+
+# colnames(analysis_data)
+
+# model <- glm(age ~ species + sex + mother_father_same + month_born + litter_size, family = gaussian(), data = analysis_data)
+
+# summary(model)
+
+wild_data <- read.csv("data/analysis_data/wild.csv")
+wild_data <- wild_data %>% drop_na(sex, species, age, month_born)
+# give count of na valus in each column
+# Extract function to count NA values in each column
+# Count NA values in each column of wild_data
+count_na(wild_data)
+# drop na values in columns sex, species, age, month_born
+# create a model for wild animals
+wild_model <- glm(age ~ species + sex + month_born, family = gaussian(), data = wild_data)
+summary(wild_model)
+
+captive_data <- read.csv("data/analysis_data/captive.csv")
+# drop na values in columns sex, species, age, month_born
+captive_data <- captive_data %>% drop_na(sex, species, age, month_born)
+# create a model for captive animals
+captive_model <- glm(age ~ species + sex + month_born, family = gaussian(), data = captive_data)
+summary(captive_model)

### 2024-11-23 paper/paper.qmd
data section updated in paper and rproj file added

@@
 ---
-title: "Himalays"
+title: "Predicting Lemur's Lifespan"
 # subtitle: "My subtitle if needed"
@@
 number-sections: true
-# bibliography: references.bib
+bibliography: ref.bib
+editor:
+  render-on-save: true
 ---
-
 # Possible topics
 
-1. Predicting the average height the person will be able to climb, or the percentage of heigh depending on the various variables.
-
-# Lemur Data
+# Introduction 
+We are aiming to predict the lifepspan of a lemur.
+Specifically, using sex, specis, birth month of lemurs.
+We have also divided the dataset into two parts: one for lemurs in captivity and one for lemurs in wild.
+The goal is to realize the success we have achievd by captiving lemurs, see areas of improvement in the pre existing methods.
 
@@
 
-## Why Lemur?
+# Data
+The data was taken from @cookson_duke_lemur_center, who acquired it from Duke Lemur Center more about which, can be found in Section [3.1](#data_collection). We used @citeR
+
+<!-- TODO: Add other libraries that were used -->
+
+It is important to know about Lemurs and what affects their lifespan because they are the most endangered mammals on the planet.
+The Duke Lemur Center (DLC) is a global leader in the research, care, and conservation of lemurs.
+The DLC hosts the most diverse population of lemurs outside their native habitat in Madagascar.
+
+To do so, I chose the following variables from the dataset, which I believe are the most important factors that can affect the lifespan of a lemur:
+
+- `sex`: The sex of the lemur on birth. It is a cateogrical variable that can be `M` or `F`.
+- `species`: The species of the lemur. It is also a categorical
+
+<!-- Inclue genus from taxonomic code? -->
+
+<!-- TODO: appendix me
+- litter size par nirbhar hai
+- maa baap same par nirbhar hai?
+- agar pregnant hoti hai, toh kya age par fark aata hai?
+ -->
+
+## Data Collection {#data_collection}
+The data in these sources was acquired and processed by staff at the Duke Lemur Center (DLC).
+
+### Data Acquisition
+ As @3zehr points out, DLC staff collected data about the lemurs according to standard operating procedures and USDA, AZA, and IACUC guidelines.
+They recorded information about births, deaths, weights, enclosure moves, behaviors, and other significant events on a daily basis.
+Originally, this data was stored in handwritten and typed paper formats.
+Later, it was computerized.
+
+In the mid-1990s, the DLC started using two databases: the Animal Record Keeping System (ARKS) and MedARKS.
+These databases allowed the DLC to share information with other organizations through the International Species Information System (ISIS).
+The DLC is currently transitioning to using the Zoological Information Management System (ZIMS).
+Data not stored in these databases has been stored in spreadsheets, and the DLC is working on transferring data from older records into these databases.
+
+### Data Processing
+The DLC used SAS software to build a database for the lemur data.
+Data from various sources was imported into SAS Enterprise Guide, including ARKS, MedARKS, ZIMS, and spreadsheets.
+They wrote programs in SAS to extract, match, and join data, calculate new variables, and format the output.
+They also used tools within SAS Enterprise Guide Projects for calculations and formatting.
+The DLC uses a unique ID to match data for individual animals, and the taxonomic name for species-related variables.
+The data was validated by identifying and locating missing data, standardizing codes and text, investigating outliers, and comparing known values to the database output.Data that could not be verified was excluded from the published dataset.
+
+The DLC created two data files from the database: **the DLC Animal List**, which contains single-copy variables for each animal in the colony's history, and **the DLC Weight File**, which contains all weight measurements for each animal.
+We used the DLC Animal List for this analysis, which was first cleaned by @cookson_duke_lemur_center and then we clean it as per our needs.
+
+The data in these sources was updated on February 8, 2019. The DLC plans to update the data on a yearly basis.
+
+# Why Lemur?
 
@@
 cite(https://lemur.duke.edu)
+
+\newpage

### 2024-11-24 scripts/02-clean_data.r
scripts updated and TODOs added in ReadMe

@@ -15,10 +15,6 @@ library(tidyverse)
 ## Data Cleaning ##
 data <- read.csv("data/raw_data/animals.csv")
 
-# get unique values of taxonomic code and number of unique values
-unique(data$taxonomic_code)
-# species
-unique(data$species)
 
 # selecting relevant columns
 data <- data[ , c("animal_id", "taxonomic_code", "sex", "birth_date", "death_date", "birth_type", "litter_size", "mother_species", "father_species")]
@@ -26,34 +22,40 @@ data <- data[ , c("animal_id", "taxonomic_code", "sex", "birth_date", "death_dat
 # make first letter of taxnomic code as Genus
 # the rest of the letters as species
 
-data$species <- substr(data$taxonomic_code, 1, 1)
-data$genus <- substr(data$taxonomic_code, 2, nchar(data$taxonomic_code))
+data$genus <- substr(data$taxonomic_code, 1, 1)
+data$species <- substr(data$taxonomic_code, 2, nchar(data$taxonomic_code))
 
-
+# remove the taxonomic_code column, as it is no longer needed
+data <- data[, -2]
 
 # create a column, which says if mother and father species is same
 data$mother_father_same <- ifelse(data$mother_species == data$father_species, 1, 0)
 
+
+
 # remove the mother_species and father_species columns, as they are no longer needed
-data <- data[, -c(8, 9)]
+data <- data[, -c(7, 8)]
 
 # create a column for age
 data$birth_date <- as.Date(data$birth_date)
 data$death_date <- as.Date(data$death_date)
 data$age <- as.numeric(difftime(data$death_date, data$birth_date, units = "days"))/365
 
+
 # create a column for month born
 data$month_born <- as.numeric(format(data$birth_date, "%m"))
 
 # remove birth_date and death_date columns, as they are no longer needed
-data <- data[, -c(4, 5)]
+
+data <- data[, -c(3, 4)]
+
 # splilt the data into birth_type == captive or wild
 data_captive <- data[data$birth_type == "captive", ]
 
 data_wild <- data[data$birth_type == "wild", ]
+
 # since there is no data on the mothr father and litter size of iwld data, we remove these cols
-data_wild <- data_wild[, -c(5, 6)]
-head(data_wild)
+data_wild <- data_wild[, -c(3, 7)]
 # remove rows with missing values except missing in litter_size and mother_father_same
[…]

### 2024-11-24 scripts/04-data-analysis.r

@@ -23,19 +23,19 @@ library(tidyverse)
 # summary(model)
 
 wild_data <- read.csv("data/analysis_data/wild.csv")
-wild_data <- wild_data %>% drop_na(sex, species, age, month_born)
+
 # give count of na valus in each column
 # Extract function to count NA values in each column
 # Count NA values in each column of wild_data
 count_na(wild_data)
 # drop na values in columns sex, species, age, month_born
 # create a model for wild animals
-wild_model <- glm(age ~ species + sex + month_born, family = gaussian(), data = wild_data)
+wild_model <- glm(age ~ species + sex + month_born + genus, family = gaussian(), data = wild_data)
 summary(wild_model)
 
 captive_data <- read.csv("data/analysis_data/captive.csv")
 # drop na values in columns sex, species, age, month_born
-captive_data <- captive_data %>% drop_na(sex, species, age, month_born)
+
 # create a model for captive animals
-captive_model <- glm(age ~ species + sex + month_born, family = gaussian(), data = captive_data)
+captive_model <- glm(age ~ species + genus + sex + month_born, family = gaussian(), data = captive_data)
 summary(captive_model)

### 2024-11-24 scripts/figure.r

@@ -1,0 +1,53 @@
+# Load required libraries
+library(ggplot2)
+library(dplyr)
+
+# Read data
+wild_data <- read.csv("data/analysis_data/wild.csv")
+captive_data <- read.csv("data/analysis_data/captive.csv")
+
+# Load required libraries
+library(ggplot2)
+library(dplyr)
+library(forcats)
+
+# Add a source column to differentiate between wild and captive data
+wild_data <- wild_data %>%
+  mutate(source = "Wild")
+captive_data <- captive_data %>%
+  mutate(source = "Captive")
+
+# convert animal_id to integer
+wild_data$animal_id <- as.integer(wild_data$animal_id)
+
+# Combine datasets
+combined_data <- bind_rows(wild_data, captive_data)
+
+# Reusable plot function
+create_plot <- function(data, aes_var, fill_var, title, x_label) {
+  ggplot(data, aes(x = !!sym(aes_var), fill = !!sym(fill_var))) +
+    geom_bar(position = "dodge") +
+    labs(title = title, x = x_label, y = "Count", fill = "Source") +
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
+}
+
+# Modify data for ordering by frequency
+combined_data <- combined_data %>%
+  mutate(
+    species = fct_infreq(species),
+    genus = fct_infreq(genus),
+    month_born = fct_infreq(as.factor(month_born))
+  )
+
+# Generate combined plots
+sex_plot <- create_plot(combined_data, "sex", "source", "Sex Distribution", "Sex")
+species_plot <- create_plot(combined_data, "species", "source", "Species Distribution", "Species")
+genus_plot <- create_plot(combined_data, "genus", "source", "Genus Distribution", "Genus")
+month_plot <- create_plot(combined_data, "month_born", "source", "Birth Month Distribution", "Birth Month")
+
+# Display plots
+sex_plot
+species_plot
+genus_plot
+month_plot

### 2024-11-24 paper/paper.qmd
llm usage added and model section added to paper

@@
 bibliography: ref.bib
-editor:
-  render-on-save: true
+# editor:
+#   render-on-save: true
 ---
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+library(tidyverse)  
+library(ggplot2)
+library(dplyr)
+library(patchwork) # For arranging plots side by side
+
+# Set the color for lemurs
+WILD_COLOR <- "#006400"
+CAPTIVE_COLOR <- "#FF8C00"
+```
+
 # Possible topics
@@
 
-<!-- TODO: Add other libraries that were used -->
-
 It is important to know about Lemurs and what affects their lifespan because they are the most endangered mammals on the planet.
@@
 
-- `sex`: The sex of the lemur on birth. It is a cateogrical variable that can be `M` or `F`.
-- `species`: The species of the lemur. It is also a categorical
+- `Sex`: The sex of the lemur on birth. It is a cateogrical variable that can be `M` or `F`.
+- `Species`: The species of the lemur.
+It is also a categorical variable and can take one of the following values: `GG`, `COL`, `UL`, `RUF`, `MOH`, `MAC`, `CAT`, `FUL`, `ALB`, `AR`, `VV`, `COQ`, `MED`, `MUR`, `COU`, `TAR`, `PYG`, `ZAZ`, `MON`, `RUB`, `COR`, `SAN`, `FLA`, `MAD`, `POT`.
+The specific or common names of these species can be found in Appendix [A](#appendix_a).
+- `Genus`: The genus of the lemur. It is a categorical variable and can take one of the following values: `O`, `E`, `G`, `H`, `L`, `V`, `P`, `C`, `M`, `N`, `D`.
+- `month_born`: The month in which the lemur was born.
+It is a categorical variable and can take one of the following values: `1`, `2`, `3`, `4`, `5`, `6`, `7`, `8`, `9`, `10`, `11`, `12`, where `1` corresponds to January, `2` to February, and so on.
 
-<!-- Inclue genus from taxonomic code? -->
+Finally, we have the target variable `Age` which is the lifespan of the lemur in years. It is a continuous variable.
 
-<!-- TODO: appendix me
-- litter size par nirbhar hai
-- maa baap same par nirbhar hai?
-- agar pregnant hoti hai, toh kya age par fark aata hai?
- -->
+The following [plots](@fig-count-plots) show the distribution of the lemurs in the dataset based on the variables `Sex`, `Species`, `Genus`, and `month_born` for both lemurs in captivity and in the wild.
+
+For `Sex`, we observe that in the wild, the distribution of male and female lemurs is nearly even, with only a slight skew.
+While in captivity, there are significantly more female lemurs than males.
+The larger count of female lemurs in captivity could result from conservation breeding programs prioritizing females to ensure the species' survival.
+
+Looking at the distribution of `Species`.
+We observe that in the wild, the distribution is relatively balanced, with a few exceptions of `MAD`, `CAT` and `MED` which appear to be low in number.
+In captivity, certain species (e.g., MUR, CAT, MED) are more represented, while others are rare or absent.
+This can be due to captive environments may focus on saving the most endangered species, as we see the ones least in wild like `CAT` and `MED` are more in captivity.
+
+With the exception of genus `E`, genus representation is balanced in both wild and captive lemurs.
+`E` might be high in both cases, as it could be a common genus or have a higher conservation priority.
+
+For birth month, births in the wild are concentrated in a few months, suggesting a breeding season.
+While in captivity, births are more evenly spread across the year, with some peaks in specific months.
+Seasonal breeding in the wild is influenced by environmental factors like food availability and climate.
+Captivity can disrupt these natural cycles, as controlled environments and year-round resources allow for more frequent and less seasonal reproduction.
+
+For `Age`, we observe that in the wild, the age distribution of lemurs shows a higher frequency of younger lemurs, with relatively few individuals reaching older ages.
+In captivity, the age distribution extends further, with a significant number of lemurs living to older ages.
+This difference can be attributed to the controlled environments in captivity, which provide consistent food, medical care, and protection from predators, thus increasing longevity.
+In contrast, the challenges of the wild, such as predation, disease, and fluctuating resources, contribute to shorter lifespans.
+With there being one exception, we observe signigicant number of captive lemurs below age 1, this can be due to the extensive breeding of captivity or could be a sign of data quality issues.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-count-plots
+
+# Read data
+wild_data <- read.csv("../data/analysis_data/wild.csv")
+captive_data <- read.csv("../data/analysis_data/captive.csv")
+
+# Reusable plot function
+create_plot <- function(data, aes_var, fill_color, title, x_label) {
+  ggplot(data, aes(x = !!sym(aes_var))) +
+    geom_bar(fill = fill_color) +
+    labs(title = title, x = x_label, y = "Count") +
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
+}
+
+# Modify data for ordering by frequency
+wild_data <- wild_data %>%
+  mutate(species = fct_infreq(species),
+         genus = fct_infreq(genus),
+         month_born = fct_infreq(as.factor(month_born)))
+
+captive_data <- captive_data %>%
+  mutate(species = fct_infreq(species),
+         genus = fct_infreq(genus),
+         month_born = fct_infreq(as.factor(month_born)))
+
+# Generate plots
+wild_plot <- create_plot(wild_data, "sex", WILD_COLOR, "Wild Lemurs", "Sex")
+captive_plot <- create_plot(captive_data, "sex", CAPTIVE_COLOR, "Captive Lemurs", "Sex")
+
+wild_plot_species <- create_plot(wild_data, "species", WILD_COLOR, "Wild Lemurs", "Species")
+captive_plot_species <- create_plot(captive_data, "species", CAPTIVE_COLOR, "Captive Lemurs", "Species")
+
+wild_plot_genus <- create_plot(wild_data, "genus", WILD_COLOR, "Wild Lemurs", "Genus")
+captive_plot_genus <- create_plot(captive_data, "genus", CAPTIVE_COLOR, "Captive Lemurs", "Genus")
+
+wild_plot_month <- create_plot(wild_data, "month_born", WILD_COLOR, "Wild Lemurs", "Birth Month")
+captive_plot_month <- create_plot(captive_data, "month_born", CAPTIVE_COLOR, "Captive Lemurs", "Birth Month")
+```
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-sex-plots
+#| fig-cap: Distribution of Lemurs by Sex
+# Display plots
+wild_plot + captive_plot
+```
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-species-plots
+wild_plot_species
+captive_plot_species
+wild_plot_genus + captive_plot_genus + wild_plot_month + captive_plot_month
+```
+
+We will also look at the distribution of the target variable `Age` in the dataset.
+The following [plot](@fig-age-distribution) shows the distribution of the ages of the lemurs in the dataset for both lemurs in captivity and in the wild.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-age-distribution
+#| fig-cap: Distribution of ages of lemurs in the dataset for both lemurs in captivity and in the wild.
+
+# line plot of count of lemurs by age
+wild_plot_age <- ggplot(wild_data, aes(x = age)) +
+  geom_histogram(fill = WILD_COLOR, bins = 30) +
+  labs(title = "Wild Lemurs",
+       x = "Age",
+       y = "Count") +
+  theme_minimal()
+
+captive_plot_age <- ggplot(captive_data, aes(x = age)) +
+  geom_histogram(fill = CAPTIVE_COLOR, bins = 30) +
+  labs(title = "Captive Lemurs",
+       x = "Age",
+       y = "Count") +
+  theme_minimal()
+
+wild_plot_age + captive_plot_age
+```
 
@@
 
+# Model
+
+I used Generalized Linear Model (GLM) framework with a Gaussian family distribution to determine the expected lifespan of a lemur based on the selected variables.
+GLM is a statistical model that generalizes linear regression to include non-normal distributions.
+It is used when the dependent variable is not normally distributed or when the relationship between the dependent and independent variables is not linear.
+The Gaussian family distribution is used when the dependent variable is continuous and normally distributed, which was in my case.
+
+My model will be based on `Sex`, `Species`, `Genus`, and `month_born` as the independent variables and `Age` as the dependent variable.
+
+Mathematically, the model can be represented as:
+
+\begin{equation}
+\bar{a} = \beta_0 + \beta_1 \times \texttt{Sex} + \beta_2 \times \texttt{Species} + \beta_3 \times \texttt{Genus} + \beta_4 \times \texttt{month\_born} + \epsilon
+\end{equation}
+where,
+
+- $\bar{a}$ is the expected lifespan of a lemur.
+- $\beta_0$ is the intercept term. It represents the expected lifespan of a lemur when all other variables are zero.
+- $\beta_1$ is the coefficient for `Sex`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Sex`.
+- $\beta_2$ is the coefficient for `Species`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Species`.
+- $\beta_3$ is the coefficient for `Genus`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Genus`.
+- $\beta_4$ is the coefficient for `month_born`. It represents the change in the expected lifespan of a lemur for a one-unit change in `month_born`.
+- $\epsilon$ is the error term.
+
+# Appendix 
+\appendix
+
+# Appendix A: Species of Lemurs {#appendix_a}
+The following are the species of lemurs that are present in the dataset:
+
+- `GG`: Genus *Genus*
+
 \newpage

### 2024-11-25 paper/paper.qmd
predictions added updated and scripts cleaned.

@@
 \end{equation}
+
+$$
+\begin{aligned}
+\bar{a}_\text{wild} & \sim \mbox{Normal}(16, 4)\\
+\bar{a}_\text{captive} & \sim \mbox{Normal}(26, 4)\\
+\beta_0 & \sim \mbox{Normal}(0, 2.5)\\
+\beta_1 & \sim \mbox{Normal}(0, 2.5)\\
+\beta_2 & \sim \mbox{Normal}(0, 2.5)\\
+\beta_3 & \sim \mbox{Normal}(0, 2.5)\\
+\beta_4 & \sim \mbox{Normal}(0, 2.5)
+\end{aligned}
+$$
 where,
 
-- $\bar{a}$ is the expected lifespan of a lemur.
+- $\bar{a}_\text{wild}$ is the expected lifespan of a lemur in the wild.
+- $\bar{a}_\text{captive}$ is the expected lifespan of a lemur in captivity.
 - $\beta_0$ is the intercept term. It represents the expected lifespan of a lemur when all other variables are zero.
@@
 
+We chose a Gaussian family distribution for the model because the dependent variable `Age` is continuous and normally distributed.
+From @citenjgov we found that the expected lifespan of a lemur in wild is upto 18 years, and in captivity about 30 years.
+So, we set the prior for the expected lifespan of a lemur in the wild to be $\mbox{Normal}(16, 4)$ and in captivity to be $\mbox{Normal}(26, 4)$.
+We set the priors for the coefficients to be $\mbox{Normal}(0, 2.5)$, because we do not have any prior information about the effect of the independent variables on the dependent variable.
+
+Using moderately broad priors can help regularize the model, reducing the risk of overfitting and make us have more stable estimates, especially in scenarios with limited data.
+
+@app_model-dataset covers coefficients, standard errors, convergence checks and other model diagnostics for our model.
+
 # Appendix
+
 \appendix
 
-# Appendix A: Species of Lemurs {#appendix_a}
+# Data Cleaning {#app_data_cleaning}
+I took the following steps to clean the raw data and prepare it for analysis:
+
+1. Loading the Data: The raw data is loaded from the file data/raw_data/animals.csv.
+
+2. Column Selection: The dataset is reduced to relevant columns: `animal_id`, `taxonomic_code`, `sex`, `birth_date`, `death_date`, `birth_type`, `litter_size`, `mother_species`, and `father_species`.
+
+3. Filtering Rows: Rows where death_date is missing are removed to ensure the data includes only animals with complete lifecycle information.
+
+4.
+Extracting Genus and Species: The taxonomic_code column is split into two new columns: `genus` (first letter) and `species` (remaining characters).
+The original taxonomic_code column is removed as it is no longer needed.
+
+5.
+Age Calculation: A new column, `age`, is calculated based on the difference between `death_date` and `birth_date`, measured in years.Lemurs with an age of less than 1 year are removed to focus on mature animals and ignore infant mortality.
+This was especially common for captive-born animals because excess breeding can lead to high infant mortality rates.
+
+6. Birth Month Extraction: A new column, month_born, is created to indicate the birth month extracted from the birth_date. We needed this information to analyze seasonal effects on lifespan.
+
+7. Splitting Data by Birth Type:
+The data is split into two subsets based on birth_type: one for animals born in captivity (`data_captive`) and another for those born in the wild (`data_wild`).
+
+8. Column Removal for Wild Data: Since information on mother_species, father_species, and litter_size is unavailable for wild-born animals, these columns are removed from data_wild.
+
+9. Handling Missing Values: Both datasets, data_wild and data_captive, are cleaned by dropping rows with missing values in key columns: sex, species, age, month_born, and genus.
+
+10. Final Output: The cleaned datasets are saved as CSV files:
+Wild-born animals: `data/analysis_data/wild.csv`
+Captive-born animals: `data/analysis_data/captive.csv`
+
+This systematic cleaning ensures the dataset is consistent, reliable, and ready for further analysis.
+
+# Analysis Dataset {#app_analysis_dataset}
+
+Hello
+
+# Model Dataset {#app-model-dataset}
+
+#  Species of Lemurs {#appendix_a}
 The following are the species of lemurs that are present in the dataset:

### 2024-11-25 scripts/02-clean_data.r

@@ -10,18 +10,16 @@
 library(tidyverse)
 
 ## Load the data ##
-# REMOVE: w <- read.csv("data/raw_data/weights.csv")
-
-## Data Cleaning ##
 data <- read.csv("data/raw_data/animals.csv")
-
 
 # selecting relevant columns
 data <- data[ , c("animal_id", "taxonomic_code", "sex", "birth_date", "death_date", "birth_type", "litter_size", "mother_species", "father_species")]
 
+# remove rows where death date is not mentioned
+data <- data[!is.na(data$death_date), ]
+
 # make first letter of taxnomic code as Genus
 # the rest of the letters as species
-
 data$genus <- substr(data$taxonomic_code, 1, 1)
 data$species <- substr(data$taxonomic_code, 2, nchar(data$taxonomic_code))
 
@@ -31,8 +29,6 @@ data <- data[, -2]
 # create a column, which says if mother and father species is same
 data$mother_father_same <- ifelse(data$mother_species == data$father_species, 1, 0)
 
-
-
 # remove the mother_species and father_species columns, as they are no longer needed
 data <- data[, -c(7, 8)]
 
@@ -41,33 +37,36 @@ data$birth_date <- as.Date(data$birth_date)
 data$death_date <- as.Date(data$death_date)
 data$age <- as.numeric(difftime(data$death_date, data$birth_date, units = "days"))/365
 
+# remove lemurs with age less than 2
+data <- data[data$age >= 1, ]
+
 
 # create a column for month born
 data$month_born <- as.numeric(format(data$birth_date, "%m"))
 
 # remove birth_date and death_date columns, as they are no longer needed
-
 data <- data[, -c(3, 4)]
 
 # splilt the data into birth_type == captive or wild
 data_captive <- data[data$birth_type == "captive", ]
-
 data_wild <- data[data$birth_type == "wild", ]
 
 # since there is no data on the mothr father and litter size of iwld data, we remove these cols
 data_wild <- data_wild[, -c(3, 7)]
-# remove rows with missing values except missing in litter_size and mother_father_same
-# data <- data %>% drop_na(animal_id, species, sex, birth_type, age, month_born)
 
-# drop na
-# data <- data %>% drop_na()
-
-# save the cleaned data
+# drop relevant NA values
 data_wild <- data_wild %>% drop_na(sex, species, age, month_born, genus)
[…]

### 2024-11-25 scripts/04-data-analysis.r

@@ -1,41 +1,41 @@
 ## Preamble ##
-# Purpose: Data analysis for the project. Specifically, we will build a model to predict the age of animals based on various factors.
+# Purpose: Data analysis for the project. Specifically, we will build a model 
+#          to predict the age of animals based on various factors.
 # Author: Khushaal Nandwani
 # Date: 18 November 2024
 # Contact: khushaal.nandwani@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None
 
-count_na <- function(data) {
-    sapply(data, function(x) sum(is.na(x)))
-}
 ## Workspace setup ##
 library(tidyverse)
+library(arrow)
 
-## Modeling ##
-# Read the cleaned analysis dataset
-# analysis_data <- read.csv("data/analysis_data/analysis_data.csv")
-
-# colnames(analysis_data)
-
-# model <- glm(age ~ species + sex + mother_father_same + month_born + litter_size, family = gaussian(), data = analysis_data)
-
-# summary(model)
-
+# Load data
 wild_data <- read.csv("data/analysis_data/wild.csv")
-
-# give count of na valus in each column
-# Extract function to count NA values in each column
-# Count NA values in each column of wild_data
-count_na(wild_data)
-# drop na values in columns sex, species, age, month_born
-# create a model for wild animals
-wild_model <- glm(age ~ species + sex + month_born + genus, family = gaussian(), data = wild_data)
-summary(wild_model)
-
 captive_data <- read.csv("data/analysis_data/captive.csv")
-# drop na values in columns sex, species, age, month_born
 
-# create a model for captive animals
-captive_model <- glm(age ~ species + genus + sex + month_born, family = gaussian(), data = captive_data)
-summary(captive_model)
+# Build models
+wild_model <- 
+    glm(
+        age ~ species + sex + month_born + genus, 
+        family = gaussian(), 
+        data = wild_data,
+        adapt_delta = 0.95,
+        prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+        prior_intercept = normal(location = 16, scale = 4, autoscale = TRUE)
+    )
+captive_model <- 
+    glm(
+        age ~ species + genus + sex + month_born, 
+        family = gaussian(), 
+        data = captive_data,
+        adapt_delta = 0.95,
+        prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
[…]

### 2024-11-25 scripts/05-predicting.r

@@ -1,0 +1,57 @@
+#### Preamble ####
+# Purpose: Creates prediction dataset to use in the Results section of the paper
+# Author: Kaavya Kalani
+# Date: 18 April 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: Run 04-data-analysis.r
+
+## Workspace setup ##
+library(tidyverse)
+library(arrow)
+
+# Prediction datasets
+wild_analysis_data <- read_csv("data/analysis_data/wild.csv")
+captive_analysis_data <- read_csv("data/analysis_data/captive.csv")
+
+# Load models
+wild_model <- readRDS("models/wild_model.rds")
+captive_model <- readRDS("models/captive_model.rds")
+
+# creating prediction dataset
+wild_ages_ <- unique(wild_analysis_data$age)
+wild_species_ <- unique(wild_analysis_data$species)
+wild_sex_ <- unique(wild_analysis_data$sex)
+wild_month_born_ <- unique(wild_analysis_data$month_born)
+wild_genus_ <- unique(wild_analysis_data$genus)
+
+captive_ages_ <- unique(captive_analysis_data$age)
+captive_species_ <- unique(captive_analysis_data$species)
+captive_sex_ <- unique(captive_analysis_data$sex)
+captive_month_born_ <- unique(captive_analysis_data$month_born)
+captive_genus_ <- unique(captive_analysis_data$genus)
+
+wild_prediction_data <- expand.grid(
+    age = wild_ages_,
+    species = wild_species_,
+    sex = wild_sex_,
+    month_born = wild_month_born_,
+    genus = wild_genus_
+)
+
+captive_prediction_data <- expand.grid(
+    age = captive_ages_,
+    species = captive_species_,
+    sex = captive_sex_,
+    month_born = captive_month_born_,
+    genus = captive_genus_
+)
+
+# predict age based on these variables
+predictions_wild <- predict(wild_model, newdata = wild_prediction_data, type = "response")
+predictions_captive <- predict(captive_model, newdata = captive_prediction_data, type = "response")
+
+
+# Save the predictions
+write.csv(predictions_wild, "data/predictions/wild_predictions.csv", row.names = FALSE)
+write.csv(predictions_captive, "data/predictions/captive_predictions.csv", row.names = FALSE)

### 2024-11-26 scripts/04-data-analysis.r
analysis model priors added and changed to brms

@@ -11,31 +11,45 @@
 library(tidyverse)
 library(arrow)
 
+library(brms)
 # Load data
 wild_data <- read.csv("data/analysis_data/wild.csv")
 captive_data <- read.csv("data/analysis_data/captive.csv")
 
 # Build models
-wild_model <- 
-    glm(
-        age ~ species + sex + month_born + genus, 
-        family = gaussian(), 
-        data = wild_data,
-        adapt_delta = 0.95,
-        prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-        prior_intercept = normal(location = 16, scale = 4, autoscale = TRUE)
-    )
-captive_model <- 
-    glm(
-        age ~ species + genus + sex + month_born, 
-        family = gaussian(), 
-        data = captive_data,
-        adapt_delta = 0.95,
-        prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-        prior_intercept = normal(location = 26, scale = 4, autoscale = TRUE)
-    )
+wild_model <- brm(
+    formula = age ~ species + sex + month_born + genus,
+    family = gaussian(),
+    data = wild_data,
+    prior = c(
+        set_prior("normal(0, 2.5)", class = "b"),
+        set_prior("normal(16, 4)", class = "Intercept")
+    ),
+    control = list(adapt_delta = 0.95)
+)
+
+# captive_model <- 
+#     glm(
+#         age ~ species + genus + sex + month_born, 
+#         family = gaussian(), 
+#         data = captive_data,
+#         prior = c(
+#             set_prior("normal(0, 2.5)", class = "b"),
+#             set_prior("normal(26, 4)", class = "Intercept")
+#         ),
+#         control = list(adapt_delta = 0.95)
+#     )
+captive_model <- brm(
+    formula = age ~ species + sex + month_born + genus,
+    family = gaussian(),
+    data = captive_data,
+    prior = c(
+        set_prior("normal(0, 2.5)", class = "b"),
+        set_prior("normal(26, 4)", class = "Intercept")
+    ),
+    control = list(adapt_delta = 0.95)
+)
 
 # save models
 saveRDS(wild_model, "models/wild_model.rds")
 saveRDS(captive_model, "models/captive_model.rds")
-

### 2024-11-27 paper/paper.qmd
results section completed

@@
 CAPTIVE_COLOR <- "#FF8C00"
+
+predictions <- read_csv("../data/predictions/predictions.csv")
 ```
@@
 
+# Results
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-wild-model-coefficients
+#| fig-cap: Coefficients of the wild model
+
+wild_model <- readRDS("../models/wild_model.rds")
+
+# Coefficients for the wild model
+coefficients <- broom::tidy(wild_model, conf.int = TRUE)
+# remove NAs
+coefficients <- coefficients %>% drop_na()
+# remove the ones which are not significant
+coefficients <- coefficients %>%
+  filter(
+    (str_starts(term, "sex")) | 
+    p.value <= 0.07
+  )
+
+# Plot for the coefficients of the model
+ggplot(coefficients, aes(x = estimate, y = term)) +
+  geom_point() + # Coefficient points
+  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) + # Confidence/Credible interval bars
+  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # Add zero line
+  labs(x = "Coefficient Estimate", y = "Predictor Variable") + # Axis labels
+  theme_minimal() # Minimal theme for better readability
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-captive-model-coefficients
+#| fig-cap: Coefficients of the captive model
+
+captive_model <- readRDS("../models/captive_model.rds")
+
+# Coefficients for the captive model
+coefficients <- broom::tidy(captive_model, conf.int = TRUE)
+
+# remove NAs
+coefficients <- coefficients %>% drop_na()
+# remove the ones which are not significant
+coefficients <- coefficients %>%
+  filter(
+    (str_starts(term, "sex") | str_starts(term, "month_born")) | 
+    p.value <= 0.05
+  )
+
+# Plot for the coefficients of the model
+ggplot(coefficients, aes(x = estimate, y = term)) +
+  geom_point() + # Coefficient points
+  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) + # Confidence/Credible interval bars
+  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # Add zero line
+  labs(x = "Coefficient Estimate", y = "Predictor Variable") + # Axis labels
+  theme_minimal() # Minimal theme for better readability
+```
+
+@fig-wild-model-coefficients and @fig-captive-model-coefficients shows the coefficients and their 95% credible intervals for the predictor variables in the respective wild and captive generalized linear regression models.
+A 95% credible interval indicates a 95% chance that the true parameter value falls within the interval, based on the observed data and model assumptions.
+Each point represents the estimated coefficient for a predictor, and the horizontal lines show the credible interval around it.
+Coefficients to the right of zero suggest a positive relationship with the outcome variable, meaning that as the predictor increases, the outcome also increases.
+Coefficients to the left of zero suggest a negative relationship, meaning that as the predictor increases, the outcome decreases.
+
+## Wild Lemurs
+Few species in the wild have significant effect on the age, these include `VV`, `SAN`, `RUF`, `RUB`, `MAD` and `CAT`.
+For sex, as expected in wild, males have a higher lifespan than females.
+We had to jump over p-value of 0.07 to get the significant results, as we had limited data in case of wild.
+
+## Captive Lemurs
+Few species have a negative effect on age, these include `MUR`, `MOH`, `MAD`, `GG`, `COQ` and `AR`.
+In case of captivity, sex is not a major factor in age, as its confidence interval is nearly centered at 0.
+Overall, species differences are prominent, while birth month effects appear less consistent, mostly centered around 0.
+Except December and September seem to be negatively affecting the age of lemurs in captivity, while November positively.
+
+## Effect of Sex on Age
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-sex-effects
+#| fig-cap: Comparison of Average Predicted Ages of Lemurs by Age Range, Sex, and Prediction Type (Wild vs. Captive)
+
+# Define age ranges
+predictions <- predictions %>%
+  mutate(age_range = cut(age, breaks = c(0, 10, 20, 40), labels = c("0-10", "10-20", "20+"), include.lowest = TRUE))
+
+# Summarize data for plotting
+plot_data <- predictions %>%
+  group_by(age_range, sex) %>%
+  summarise(
+    avg_age_wild = mean(predicted_age_wild, na.rm = TRUE),
+    avg_age_captive = mean(predicted_age_captive, na.rm = TRUE)
+  ) %>%
+  pivot_longer(
+    cols = c(avg_age_wild, avg_age_captive),
+    names_to = "prediction_type",
+    values_to = "avg_age"
+  )
+
+# remove NAs
+plot_data <- plot_data %>% drop_na()
+
+# Create the bar plot
+plotted <- ggplot(plot_data, aes(x = age_range, y = avg_age, fill = prediction_type)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  facet_wrap(~sex) +
+  scale_fill_manual(
+    values = c("avg_age_wild" = WILD_COLOR, "avg_age_captive" = CAPTIVE_COLOR),
+    labels = c("Wild Lemurs", "Captive Lemurs")
+  ) +
+  labs(
+    x = "Age Ranges",
+    y = "Average Predicted Age",
+    fill = "Type of Lemurs"
+  ) +
+  theme_minimal()
+
+plotted
+```
+
+From the @fig-sex-effects, lemurs in captive have a higher average predicted age than those in the wild across all age ranges.
+And in the captive, we see males living longer on average than females, while in the wild, the difference is less pronounced.
+This can be due to the controlled environment in captivity, which provides consistent food, medical care, and protection from predators, thus increasing longevity.
+In the wild, the challenges of predation, disease, and fluctuating resources contribute to shorter lifespans.
+
+## Effect of Month Born on Age
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-month-born-effects
+#| fig-cap: Comparison of Average Predicted Ages of Lemurs by Age Range, Birth Month, and Prediction Type (Wild vs. Captive)
+
+# Summarize the data by month
+month_plot_data <- predictions %>%
+  group_by(month_born) %>%
+  summarise(
+    avg_age_wild = mean(predicted_age_wild, na.rm = TRUE),
+    avg_age_captive = mean(predicted_age_captive, na.rm = TRUE)
+  ) %>%
+  pivot_longer(
+    cols = c(avg_age_wild, avg_age_captive),
+    names_to = "prediction_type",
+    values_to = "avg_age"
+  )
+
+# Create the bar plot
+month_plot <- ggplot(month_plot_data, aes(x = month_born, y = avg_age, fill = prediction_type)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  scale_fill_manual(
+    values = c("avg_age_wild" = WILD_COLOR, "avg_age_captive" = CAPTIVE_COLOR),
+    labels = c("Wild Lemurs", "Captive Lemurs")
+  ) +
+  labs(
+    x = "Month Born",
+    y = "Average Predicted Age",
+    fill = "Type of Lemurs",
+    title = "Average Predicted Ages by Month Born"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+month_plot
+```
+
+For months in @fig-month-born-effects, we see captive lemurs outlive wild lemurs most of the months except August, January, March, Novemeber and September.
+It is worth exploring the reasons behind these differences by Duke Lemur Center, as they can provide insights into the impact of seasonal factors on lemur lifespan.
+
+## Effect of Species on Age
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-species-effects
+#| fig-cap: Comparison of Average Predicted Ages of Lemurs by Age Range, Species, and Prediction Type (Wild vs. Captive)
+
+# Summarize the data by species
+species_plot_data <- predictions %>%
+  group_by(species) %>%
+  summarise(
+    avg_age_wild = mean(predicted_age_wild, na.rm = TRUE),
+    avg_age_captive = mean(predicted_age_captive, na.rm = TRUE)
+  ) %>%
+  pivot_longer(
+    cols = c(avg_age_wild, avg_age_captive),
+    names_to = "prediction_type",
+    values_to = "avg_age"
+  )
+
+# Create the bar plot
+species_plot <- ggplot(species_plot_data, aes(x = species, y = avg_age, fill = prediction_type)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  scale_fill_manual(
+    values = c("avg_age_wild" = WILD_COLOR, "avg_age_captive" = CAPTIVE_COLOR),
+    labels = c("Wild Lemurs", "Captive Lemurs")
+  ) +
+  labs(
+    x = "Species",
+    y = "Average Predicted Age",
+    fill = "Type of Lemurs",
+    title = "Average Predicted Ages by Species"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+species_plot
+```
+
+Once again in species as seen in @fig-species-effects, we see captive lemurs living longer than wild ones in most of them, except `ALB`, `COR`, `MAC` and `MON`.
+Why these species are living longer in wild, can be a good area of research for Duke Lemur Center.
+
 # Appendix
@@
 # Analysis Dataset {#app_analysis_dataset}
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-wild
+#| tbl-cap: Analysis dataset for Wild Lemurs
+library(knitr)
 
-Hello
+wild_data <- read.csv("../data/analysis_data/wild.csv")
+captive_data <- read.csv("../data/analysis_data/captive.csv")
+
+# Glimpse of the anaysis dataset
+head(wild_data) %>%
+  kable(digits = 2)
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-captive
+#| tbl-cap: Analysis dataset for Captive Lemurs
+
+head(captive_data) %>%
+  kable(digits = 2)
+```
+@tbl-wild provides a glimpse of the dataset used for analysis for wild lemurs, and @tbl-captive provides a glimpse of the dataset used for analysis for captive lemurs.
 

### 2024-11-27 scripts/02-clean_data.r
models and scripts updated

@@ -43,6 +43,9 @@ data <- data[data$age >= 1, ]
 
 # create a column for month born
 data$month_born <- as.numeric(format(data$birth_date, "%m"))
+# convert month_born to name of the months
+data$month_born <- month.abb[data$month_born]
+
 
 # remove birth_date and death_date columns, as they are no longer needed
 data <- data[, -c(3, 4)]
@@ -56,18 +59,9 @@ data_wild <- data_wild[, -c(3, 7)]
 
 # drop relevant NA values
 data_wild <- data_wild %>% drop_na(sex, species, age, month_born, genus)
+# remove litter size
+data_wild <- data_wild[, -3]
 write.csv(data_wild, "data/analysis_data/wild.csv", row.names = FALSE)
 data_captive <- data_captive %>% drop_na(sex, species, age, month_born, genus)
 write.csv(data_captive, "data/analysis_data/captive.csv", row.names = FALSE)
-
-
-
-
-
-
-
-
-
-
-
 

### 2024-11-27 scripts/04-data-analysis.r

@@ -10,46 +10,25 @@
 ## Workspace setup ##
 library(tidyverse)
 library(arrow)
-
 library(brms)
+library(rstanarm)
+
 # Load data
 wild_data <- read.csv("data/analysis_data/wild.csv")
 captive_data <- read.csv("data/analysis_data/captive.csv")
 
-# Build models
-wild_model <- brm(
+wild_model <- glm(
     formula = age ~ species + sex + month_born + genus,
     family = gaussian(),
     data = wild_data,
-    prior = c(
-        set_prior("normal(0, 2.5)", class = "b"),
-        set_prior("normal(16, 4)", class = "Intercept")
-    ),
-    control = list(adapt_delta = 0.95)
 )
-
-# captive_model <- 
-#     glm(
-#         age ~ species + genus + sex + month_born, 
-#         family = gaussian(), 
-#         data = captive_data,
-#         prior = c(
-#             set_prior("normal(0, 2.5)", class = "b"),
-#             set_prior("normal(26, 4)", class = "Intercept")
-#         ),
-#         control = list(adapt_delta = 0.95)
-#     )
-captive_model <- brm(
+captive_model <- glm(
     formula = age ~ species + sex + month_born + genus,
     family = gaussian(),
     data = captive_data,
-    prior = c(
-        set_prior("normal(0, 2.5)", class = "b"),
-        set_prior("normal(26, 4)", class = "Intercept")
-    ),
-    control = list(adapt_delta = 0.95)
 )
 
 # save models
 saveRDS(wild_model, "models/wild_model.rds")
 saveRDS(captive_model, "models/captive_model.rds")
+

### 2024-11-27 scripts/05-predicting.r

@@ -11,47 +11,41 @@ library(tidyverse)
 library(arrow)
 
 # Prediction datasets
-wild_analysis_data <- read_csv("data/analysis_data/wild.csv")
-captive_analysis_data <- read_csv("data/analysis_data/captive.csv")
+analysis_data <- read_csv("data/analysis_data/wild.csv")
 
 # Load models
 wild_model <- readRDS("models/wild_model.rds")
 captive_model <- readRDS("models/captive_model.rds")
 
 # creating prediction dataset
-wild_ages_ <- unique(wild_analysis_data$age)
-wild_species_ <- unique(wild_analysis_data$species)
-wild_sex_ <- unique(wild_analysis_data$sex)
-wild_month_born_ <- unique(wild_analysis_data$month_born)
-wild_genus_ <- unique(wild_analysis_data$genus)
+ages_ <- unique(analysis_data$age)
+species_ <- unique(analysis_data$species)
+sex_ <- unique(analysis_data$sex)
+month_born_ <- unique(analysis_data$month_born)
+genus_ <- unique(analysis_data$genus)
 
-captive_ages_ <- unique(captive_analysis_data$age)
-captive_species_ <- unique(captive_analysis_data$species)
-captive_sex_ <- unique(captive_analysis_data$sex)
-captive_month_born_ <- unique(captive_analysis_data$month_born)
-captive_genus_ <- unique(captive_analysis_data$genus)
-
-wild_prediction_data <- expand.grid(
-    age = wild_ages_,
-    species = wild_species_,
-    sex = wild_sex_,
-    month_born = wild_month_born_,
-    genus = wild_genus_
-)
-
-captive_prediction_data <- expand.grid(
-    age = captive_ages_,
-    species = captive_species_,
-    sex = captive_sex_,
-    month_born = captive_month_born_,
-    genus = captive_genus_
+prediction_data <- expand.grid(
+    age = ages_,
+    species = species_,
+    sex = sex_,
+    month_born = month_born_,
+    genus = genus_
 )
 
 # predict age based on these variables
-predictions_wild <- predict(wild_model, newdata = wild_prediction_data, type = "response")
-predictions_captive <- predict(captive_model, newdata = captive_prediction_data, type = "response")
+predictions_wild <- predict(wild_model, newdata = prediction_data, type = "response")
+predictions_captive <- predict(captive_model, newdata = prediction_data, type = "response")
 
 
-# Save the predictions
[…]

### 2024-11-27 scripts/figure.r

@@ -1,53 +1,40 @@
-# Load required libraries
 library(ggplot2)
 library(dplyr)
+WILD_COLOR <- "#FFA500"
+CAPTIVE_COLOR <- "#4169E1"
 
-# Read data
-wild_data <- read.csv("data/analysis_data/wild.csv")
-captive_data <- read.csv("data/analysis_data/captive.csv")
-
-# Load required libraries
-library(ggplot2)
-library(dplyr)
-library(forcats)
-
-# Add a source column to differentiate between wild and captive data
-wild_data <- wild_data %>%
-  mutate(source = "Wild")
-captive_data <- captive_data %>%
-  mutate(source = "Captive")
-
-# convert animal_id to integer
-wild_data$animal_id <- as.integer(wild_data$animal_id)
-
-# Combine datasets
-combined_data <- bind_rows(wild_data, captive_data)
-
-# Reusable plot function
-create_plot <- function(data, aes_var, fill_var, title, x_label) {
-  ggplot(data, aes(x = !!sym(aes_var), fill = !!sym(fill_var))) +
-    geom_bar(position = "dodge") +
-    labs(title = title, x = x_label, y = "Count", fill = "Source") +
-    theme_minimal() +
-    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
-}
-
-# Modify data for ordering by frequency
-combined_data <- combined_data %>%
-  mutate(
-    species = fct_infreq(species),
-    genus = fct_infreq(genus),
-    month_born = fct_infreq(as.factor(month_born))
+# Load the data
+predictions <- read_csv("data/predictions/predictions.csv")
+# Summarize the data by month
+month_plot_data <- predictions %>%
+  group_by(month_born) %>%
+  summarise(
+    avg_age_wild = mean(predicted_age_wild, na.rm = TRUE),
+    avg_age_captive = mean(predicted_age_captive, na.rm = TRUE)
+  ) %>%
+  pivot_longer(
+    cols = c(avg_age_wild, avg_age_captive),
+    names_to = "prediction_type",
+    values_to = "avg_age"
   )
 
-# Generate combined plots
-sex_plot <- create_plot(combined_data, "sex", "source", "Sex Distribution", "Sex")
-species_plot <- create_plot(combined_data, "species", "source", "Species Distribution", "Species")
[…]

### 2024-11-27 paper/paper.qmd
Readme updated and abstract and title added

@@
 ---
-title: "Predicting Lemur's Lifespan"
+title: "Predicting a Lemur's Lifespan in Wild and Captivity"
 # subtitle: "My subtitle if needed"
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This study aims to predict the lifespan of lemurs in wild and captive by analyzing key factors such as sex, species, genus, and birth month.
+Utilizing data from the Duke Lemur Center, I divided the dataset into wild-born and captive-born lemurs to assess differences in longevity.
+I employed Generalized Linear Models (GLMs) with a Gaussian distribution to model lifespan based on the selected variables.
+My findings indicate that lemurs in captivity have a significantly higher average lifespan compared to their wild counterparts, likely due to controlled environments and access to medical care.
+In wild populations, sex was a significant predictor of lifespan, with males living longer than females, whereas in captivity, sex did not have a substantial impact.
+Species differences were prominent in both settings, suggesting inherent biological factors affecting longevity.
+Birth month showed inconsistent effects on lifespan across both groups.
+These results highlight the influence of environmental conditions on lemur longevity and underscore the importance of species-specific conservation strategies."
 format: pdf
 number-sections: true
+toc: true
 bibliography: ref.bib

### 2024-11-27 paper/paper.qmd
discussion section added, intro updated

@@
 
-# Possible topics
+# Introduction
 
-# Introduction 
-We are aiming to predict the lifepspan of a lemur.
-Specifically, using sex, specis, birth month of lemurs.
-We have also divided the dataset into two parts: one for lemurs in captivity and one for lemurs in wild.
-The goal is to realize the success we have achievd by captiving lemurs, see areas of improvement in the pre existing methods.
+Lemurs are the world’s most endangered group of mammals (@citedukelemurcenter).
+Understanding the factors that influence their lifespan is crucial for guiding effective conservation efforts.
+The Duke Lemur Center (DLC), established in 1966, has played a pivotal role in the study and preservation of lemurs, maintaining the largest population of these species outside their native Madagascar.
+Different factors can affect the longevity of lemurs in wild and captive, longevity is important because the longer lemurs live, the more likely they are to reproduce and contribute to the survival of their species.
 
-1. What factors can contribute to lifespan of lemur?
-    - sex
-    - species
-    - birth month (season): some seasons makes stronger. maybe winter?
-    - captivity vs wild: captivity makes them use to spoon feeding?
-    - litter size: maybe mother is more involved with other kids
-    - mother_species and father_species is same or not
+This study uses a dataset from the DLC to analyze if and how sex, species, genus and month a lemur is born in can affect its lifespan in the wild and captivity.
+This study can help us guide what kinds of lemur are we failing to conserve in captivity and how they perform in wild.
+It will also reinforce where our efforts have been successful.
 
-# Data
-The data was taken from @cookson_duke_lemur_center, who acquired it from Duke Lemur Center more about which, can be found in Section [3.1](#data_collection). We used @citeR
+Using Generalized Linear Models (GLMs) with Gaussian distribution, factors including sex, species, genus, and birth month were evaluated to predict lemur longevity.
+The dataset was divided into wild-born and captive-born lemurs to understand the differences in lifespan between the two environments and separate models were built for each group.
+
+The findings reveal that lemurs in captivity exhibit significantly longer lifespans.
+In wild populations, males tend to live longer than females, while species differences are prominent in both environments, emphasizing the need for species-specific conservation strategies.
+Birth month was found to have inconsistent effects across both groups, suggesting complex interactions with environmental conditions.
+
+The remainder of this paper is structured as follows: @sec-data details the data collection and processing methods, while @sec-model outlines the modeling approach.
+@sec-results presents the results, highlighting key differences between wild and captive lemurs.
+The discussion in @sec-sec-discussion contextualizes these findings, addressing limitations and proposing future directions for research.
+Finally, the @sec-appendixs includes additional information on data cleaning, the analysis dataset, and the model dataset.
+
+# Data {#sec-data}
+The data was taken from @cookson_duke_lemur_center, who acquired it from Duke Lemur Center more about which, can be found in @sec-data_collection. We used @citeR
 
@@
 
-## Data Collection {#data_collection}
+## Data Collection {#sec-data_collection}
 The data in these sources was acquired and processed by staff at the Duke Lemur Center (DLC).
@@
 
-# Why Lemur?
-
-Established in 1966 on Duke University's campus in Durham, NC, the Duke Lemur Center (DLC) is a global leader in the research, care, and conservation of lemurs, the planet's most endangered group of mammals.
-Home to over 200 animals spanning 13 species, the DLC hosts the most diverse population of lemurs outside their native habitat in Madagascar.
-cite(https://lemur.duke.edu)
-
-# Model
+# Model {#sec-model}
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Appendix
+# Discussion {#sec-discussion}
+
+In this study, we sought to predict and analyze the lifespan of lemurs in wild and captive environments by examining factors of sex, species, genus, and birth month.
+Utilizing data from the Duke Lemur Center, we divided the dataset into wild-born and captive-born lemurs to understand how environmental conditions impact longevity.
+Generalized Linear Models (GLMs) with a Gaussian distribution were employed to understand the relationships between these variables and lifespan.
+
+## Key Findings
+
+We learn from this study that captivity plays a major role in the lifespan of lemurs.
+Captive lemurs, benefit from controlled environments, consistent food supplies, and medical care, live considerably longer on average than their wild counterparts similar to as the study by @citetidi showed.
+This underscores the potential of captive breeding programs to contribute positively to species survival and highlights the role of human intervention in extending the lifespans of endangered species.
+
+Another important finding is the variability in lifespan based on species and sex.
+In the wild, male lemurs tend to live longer than females, possibly due to biological or behavioral factors.
+However, this sex-based difference diminishes in captivity, suggesting that controlled environments mitigate sex-specific survival challenges.
+Additionally, species differences in longevity were significant in both settings, highlighting the influence of inherent biological factors and emphasizing the need for species-specific conservation strategies.
+
+## Weaknesses and Limitations
+
+Despite these findings, this study has several limitations.
+The dataset used was relatively small for wild-born lemurs, potentially affecting the reliability of the statistical models.
+For instance, the limited sample size may have led to the inclusion of predictors with p-values slightly above the conventional threshold, which could inflate Type I error rates.
+Furthermore, the data on wild-born lemurs lacked detailed information on maternal and paternal species and litter size, which are potentially critical factors in understanding lifespan dynamics, and we could not include them in our analysis.
+
+Another limitation lies in the potential bias introduced by data collection processes.
+The reliance on historical records and the transition between database systems may have resulted in inconsistencies or inaccuracies.
+Additionally, the study only considered certain categorical variables, excluding other possible factors such as habitat quality, diet variability, or exposure to diseases, which could provide a more holistic understanding of lemur longevity.
+
+## Future Directions
+
+This research opens the door to several avenues for further investigation.
+Future studies should aim to collect larger and more representative datasets for wild-born lemurs to improve the robustness of the models.
+Additionally, incorporating environmental variables such as habitat degradation, climate patterns, and predation risks could provide better understanding of the challenges faced by wild populations.
+
+Furthermore, understanding why certain species thrive better in the wild (e.g., ALB, COR, MAC, and MON) compared to captivity requires focused investigation.
+This could involve studying behavioral adaptations, ecological niches, and species-specific physiological traits.
+Expanding the analysis to include genetic data may also uncover underlying biological factors that contribute to these differences.
+
+Finally, longitudinal studies examining the long-term impacts of captivity on lemur behavior, reproduction, and overall health would be invaluable.
+Integrating interdisciplinary approaches, combining ecological, genetic, and conservation sciences, could help develop more effective strategies for preserving these critically endangered species.
+
+In conclusion, while this study provides meaningful insights into the factors influencing lemur lifespan, it also highlights the complexities of their survival in diverse environments.
+Addressing the limitations and expanding the scope of research will not only deepen our understanding but also support more informed conservation efforts for lemurs and other endangered species.
+
+# Appendix {#sec-appendix}
 
@@
 
-# Data Cleaning {#app_data_cleaning}
+# Data Cleaning {#sec-appendix-data-cleaning}
 I took the following steps to clean the raw data and prepare it for analysis:
@@
 
-# Analysis Dataset {#app_analysis_dataset}
+# Analysis Dataset {#sec-appendix-data-analysis}
 ```{r}
@@
 
-#  Species of Lemurs {#appendix_a}
+#  Species of Lemurs {#sec-appendix-species}
 The following are the species of lemurs that are present in the dataset:
 
-- `GG`: Genus *Genus*
+| Index | Code | Latin Name | Common Name |
+|-------|------|------------|-------------|
+| 1     | MED | Cheirogaleus medius | Fat-tailed dwarf lemur |
+| 2     | MAD | Daubentonia madagascariensis | Aye-aye |
+| 3     | ALB | Eulemur albifrons | White-fronted brown lemur |
+| 4     | COL | Eulemur collaris | Collared brown lemur |
+| 5     | COR | Eulemur coronatus | Crowned lemur |
+| 6     | FLA | Eulemur flavifrons | Blue-eyed black lemur |
+| 7     | FUL | Eulemur fulvus | Common brown lemur |
+| 8     | MAC | Eulemur macaco | Black lemur |
+| 9     | MON | Eulemur mongoz | Mongoose lemur |
+| 10    | RUB | Eulemur rubriventer | Red-bellied lemur |
+| 11    | RUF | Eulemur rufus | Red-fronted brown lemur |
+| 12    | SAN | Eulemur sanfordi | Sanford's brown lemur |
+| 13    | MOH | Galago moholi | Mohol bushbaby |
+| 14    | GG  | Hapalemur griseus griseus | Eastern lesser bamboo lemur |
+| 15    | CAT | Lemur catta | Ring-tailed lemur |
+| 16    | TAR | Loris tardigradus | Slender loris |
+| 17    | MUR | Mircocebus murinus | Gray mouse lemur |
+| 18    | COQ | Mirza coquereli | Northern giant mouse lemur |
+| 19    | COU | Nycticebus coucang | Slow loris |
+| 20    | PYG | Nycticebus pygmaeus | Pygmy slow loris |
+| 21    | COQ | Propithecus coquereli | Coquerel's sifaka |
+| 22    | POT | Perodicticus potto | Potto |
+| 23    | VAR | Varecia | Varecia hybrid |
+| 24    | RUB | Varecia rubra | Red ruffed lemur |
+| 25    | VV  | Varecia variegata variegata | Black-and-white ruffed lemur |
 

### 2024-11-27 paper/paper.qmd
download data and tests added

@@
   - Khushaal Nandwani
-thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
+thanks: "Code and data are available at: [https://github.com/khushaal-nandwani/predicting-lemurs/tree/main](https://github.com/khushaal-nandwani/predicting-lemurs/tree/main)."
 date: today
 date-format: long
-abstract: "This study aims to predict the lifespan of lemurs in wild and captive by analyzing key factors such as sex, species, genus, and birth month.
-Utilizing data from the Duke Lemur Center, I divided the dataset into wild-born and captive-born lemurs to assess differences in longevity.
-I employed Generalized Linear Models (GLMs) with a Gaussian distribution to model lifespan based on the selected variables.
-My findings indicate that lemurs in captivity have a significantly higher average lifespan compared to their wild counterparts, likely due to controlled environments and access to medical care.
-In wild populations, sex was a significant predictor of lifespan, with males living longer than females, whereas in captivity, sex did not have a substantial impact.
-Species differences were prominent in both settings, suggesting inherent biological factors affecting longevity.
-Birth month showed inconsistent effects on lifespan across both groups.
-These results highlight the influence of environmental conditions on lemur longevity and underscore the importance of species-specific conservation strategies."
+abstract: "This study investigates the lifespan of lemurs in both wild and captive environments using data from the Duke Lemur Center, aiming to identify how factors such as sex, species, genus, and birth month influence longevity.
+Using Generalized Linear Models, the analysis highlights that lemurs in captivity live significantly longer than their wild counterparts, emphasizing the impact of conservation practices.
+The findings reveal species-specific variations in lifespan and the diminishing influence of sex differences in captivity.
+These insights help identify effective practices in captivity and suggest areas for improvement to optimize conservation strategies, contributing to the survival of this endangered mammalian group."
 format: pdf
@@
 The data was taken from @cookson_duke_lemur_center, who acquired it from Duke Lemur Center more about which, can be found in @sec-data_collection. We used @citeR 
+tidyverse
+rstanarm
+arrow
+ggplot2
+knitr
+dplyr
 
@@
 
-2. Column Selection: The dataset is reduced to relevant columns: `animal_id`, `taxonomic_code`, `sex`, `birth_date`, `death_date`, `birth_type`, `litter_size`, `mother_species`, and `father_species`.
+2. Column Selection: The dataset is reduced to relevant columns: `animal_id`, `taxonomic_code`, `sex`, `birth_date`, `death_date` and `birth_type`.
 
@@
 5.
-Age Calculation: A new column, `age`, is calculated based on the difference between `death_date` and `birth_date`, measured in years.Lemurs with an age of less than 1 year are removed to focus on mature animals and ignore infant mortality.
+Age Calculation: A new column, `age`, is calculated based on the difference between `death_date` and `birth_date`, measured in years.
+Lemurs with an age of less than 1 year are removed to focus on mature animals and ignore infant mortality.
 This was especially common for captive-born animals because excess breeding can lead to high infant mortality rates.
 
-6. Birth Month Extraction: A new column, month_born, is created to indicate the birth month extracted from the birth_date. We needed this information to analyze seasonal effects on lifespan.
+6. Birth Month Extraction: A new column, `month_born`, is created to indicate the birth month extracted from the `birth_date`. We needed this information to analyze seasonal effects on lifespan.
 
@@
 
-8. Column Removal for Wild Data: Since information on mother_species, father_species, and litter_size is unavailable for wild-born animals, these columns are removed from data_wild.
-
-9. Handling Missing Values: Both datasets, data_wild and data_captive, are cleaned by dropping rows with missing values in key columns: sex, species, age, month_born, and genus.
-
-10. Final Output: The cleaned datasets are saved as CSV files:
+8. Final Output: NA values are then dropped and the cleaned datasets are saved as CSV files:
 Wild-born animals: `data/analysis_data/wild.csv`
@@
 
-# Model Dataset {#app-model-dataset}
-
 #  Species of Lemurs {#sec-appendix-species}

### 2024-11-27 scripts/00-install_packages.r

@@ -12,6 +12,8 @@ packages <- c(
   "dplyr",
   "knitr",
   "ggplot2",
+  "arrow",
+  "rstanarm"
 )
 
 # Install packages if they are not already installed

### 2024-11-27 scripts/02-download_data.r

@@ -1,0 +1,19 @@
+url <- "https://raw.githubusercontent.com/tacookson/data/master/duke-lemur-center/animals.txt"
+
+destination_dir <- "data/raw_data"
+destination_file <- file.path(destination_dir, "animals.txt")
+
+# Create the destination directory if it doesn't exist
+if (!dir.exists(destination_dir)) {
+  dir.create(destination_dir, recursive = TRUE)
+}
+
+# Download the file
+download.file(url, destination_file, mode = "wb")
+
+# Confirm success
+if (file.exists(destination_file)) {
+  cat("File downloaded and saved to", destination_file, "\n")
+} else {
+  cat("Failed to download the file.\n")
+}

### 2024-11-27 scripts/03-clean_data.r

@@ -10,13 +10,15 @@
 library(tidyverse)
 
 ## Load the data ##
-data <- read.csv("data/raw_data/animals.csv")
+data <- read.csv("data/raw_data/animals.txt", sep = "\t")
 
 # selecting relevant columns
-data <- data[ , c("animal_id", "taxonomic_code", "sex", "birth_date", "death_date", "birth_type", "litter_size", "mother_species", "father_species")]
+data <- data[ , c("animal_id", "taxonomic_code", "sex", "birth_date", "death_date", "birth_type")]
 
 # remove rows where death date is not mentioned
 data <- data[!is.na(data$death_date), ]
+# remove birth_type NAs
+data <- data[!is.na(data$birth_type), ]
 
 # make first letter of taxnomic code as Genus
 # the rest of the letters as species
@@ -26,26 +28,18 @@ data$species <- substr(data$taxonomic_code, 2, nchar(data$taxonomic_code))
 # remove the taxonomic_code column, as it is no longer needed
 data <- data[, -2]
 
-# create a column, which says if mother and father species is same
-data$mother_father_same <- ifelse(data$mother_species == data$father_species, 1, 0)
-
-# remove the mother_species and father_species columns, as they are no longer needed
-data <- data[, -c(7, 8)]
-
 # create a column for age
 data$birth_date <- as.Date(data$birth_date)
 data$death_date <- as.Date(data$death_date)
 data$age <- as.numeric(difftime(data$death_date, data$birth_date, units = "days"))/365
 
-# remove lemurs with age less than 2
+# remove lemurs with age less than 1
 data <- data[data$age >= 1, ]
-
 
 # create a column for month born
 data$month_born <- as.numeric(format(data$birth_date, "%m"))
 # convert month_born to name of the months
 data$month_born <- month.abb[data$month_born]
-
 
 # remove birth_date and death_date columns, as they are no longer needed
 data <- data[, -c(3, 4)]
@@ -54,14 +48,13 @@ data <- data[, -c(3, 4)]
 data_captive <- data[data$birth_type == "captive", ]
 data_wild <- data[data$birth_type == "wild", ]
 
-# since there is no data on the mothr father and litter size of iwld data, we remove these cols
-data_wild <- data_wild[, -c(3, 7)]
-
 # drop relevant NA values
 data_wild <- data_wild %>% drop_na(sex, species, age, month_born, genus)
-# remove litter size
[…]

### 2024-11-27 scripts/04-test-data.r

@@ -1,0 +1,112 @@
+#### Preamble ####
+# Purpose: Tests the analysis data
+# Author: Khushaal Nandwani
+# Date: 27 November 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: Run 03-clean_data.r
+
+## Workspace setup ##
+library(tidyverse)
+library(arrow)
+
+## Tests ##
+print("Testing analysis data for wild and captive lemurs")
+
+wild_data <- read_parquet("data/analysis_data/wild.parquet")
+captive_data <- read_parquet("data/analysis_data/captive.parquet")
+
+all_tests_passed = TRUE
+
+# Test 1: Check for any NA values or missing data
+if (any(is.na(wild_data))) {
+  print("Test 1 failed: Missing values in wild data")
+  all_tests_passed = FALSE
+}
+
+if (any(is.na(captive_data))) {
+  print("Test 1 failed: Missing values in captive data")
+  all_tests_passed = FALSE
+}
+
+# Test 2: Check number of columns
+if (ncol(wild_data) != 7) {
+  print("Test 2 failed: Incorrect number of columns in wild data")
+  all_tests_passed = FALSE
+}
+
+if (ncol(captive_data) != 7) {
+  print("Test 2 failed: Incorrect number of columns in captive data")
+  all_tests_passed = FALSE
+}
+
+# Test 3: Check age is between 0 and 50
+if (any(wild_data$age < 0) | any(wild_data$age > 50)) {
+  print("Test 3 failed: Age values out of range in wild data")
+  all_tests_passed = FALSE
+}
+
+# Test 4: Check sex is either M or F
+if (any(!(wild_data$sex %in% c("M", "F")))) {
+  print("Test 4 failed: Invalid sex in wild data")
+    all_tests_passed = FALSE
+}
+
+if (any(!(captive_data$sex %in% c("M", "F")))) {
+  print("Test 4 failed: Invalid sex in captive data")
+    all_tests_passed = FALSE
+}
+
+# Test 5: Check birth type is either captive or wild
+if (any(!(wild_data$birth_type %in% c("captive", "wild")))) {
+  print("Test 5 failed: Invalid birth type in wild data")
+  all_tests_passed = FALSE
+}
+
+if (any(!(captive_data$birth_type %in% c("captive", "wild")))) {
[…]

### 2024-11-27 scripts/05-data-analysis.r

@@ -10,12 +10,12 @@
 ## Workspace setup ##
 library(tidyverse)
 library(arrow)
-library(brms)
 library(rstanarm)
 
 # Load data
-wild_data <- read.csv("data/analysis_data/wild.csv")
-captive_data <- read.csv("data/analysis_data/captive.csv")
+wild_data <- read_parquet("data/analysis_data/wild.parquet")
+captive_data <- read_parquet("data/analysis_data/captive.parquet")
+
 
 wild_model <- glm(
     formula = age ~ species + sex + month_born + genus,
@@ -31,4 +31,3 @@ captive_model <- glm(
 # save models
 saveRDS(wild_model, "models/wild_model.rds")
 saveRDS(captive_model, "models/captive_model.rds")
-

### 2024-11-27 scripts/figure.r

@@ -1,5 +1,6 @@
 library(ggplot2)
 library(dplyr)
+
 WILD_COLOR <- "#FFA500"
 CAPTIVE_COLOR <- "#4169E1"
 

### 2024-11-28 paper/paper.qmd
refining predictions and paper prose

@@
 library(patchwork) # For arranging plots side by side
-
+library(arrow)
+library(kableExtra)
+library(rstanarm)
 # Set the color for lemurs
@@
 predictions <- read_csv("../data/predictions/predictions.csv")
+wild_model <- readRDS("../models/wild_model.rds")
+captive_model <- readRDS("../models/captive_model.rds")
 ```
@@
 # Data {#sec-data}
-The data was taken from @cookson_duke_lemur_center, who acquired it from Duke Lemur Center more about which, can be found in @sec-data_collection. We used @citeR 
-tidyverse
-rstanarm
-arrow
-ggplot2
-knitr
-dplyr
+The data was taken from @cookson_duke_lemur_center, who acquired it from Duke Lemur Center more about which, can be found in @sec-data_collection.
+We used `R` [@citeR] along with its packages to clean and analyze the data.
+These packages include `tidyverse` [@cite-tidyverse], `ggplot2` [@cite-ggplot], `dplyr` [@cite-dplyr], `patchwork` [@cite-patchwork], `arrow` [@cite-arrow], `knitr` [@cite-knitr] and `rstanarm` [@cite-rstanarm].
 
@@
 
-The following [plots](@fig-count-plots) show the distribution of the lemurs in the dataset based on the variables `Sex`, `Species`, `Genus`, and `month_born` for both lemurs in captivity and in the wild.
+The following figures show the distribution of the lemurs in the dataset based on the variables `Sex` (@fig-sex-plots), `Species` (@fig-species-effects), `Genus` (@fig-genus-plots), and `month_born` (@fig-month-born-effects) for both lemurs in captivity, shown in orange and in the wild, shown in dark green.
 
-For `Sex`, we observe that in the wild, the distribution of male and female lemurs is nearly even, with only a slight skew.
+For `Sex` (@fig-sex-plots), we observe that in the wild, the distribution of male and female lemurs is nearly even, with only a slight skew.
 While in captivity, there are significantly more female lemurs than males.
@@
 
-Looking at the distribution of `Species`.
+Looking at the distribution of `Species` (@fig-species-effects).
 We observe that in the wild, the distribution is relatively balanced, with a few exceptions of `MAD`, `CAT` and `MED` which appear to be low in number.
@@
 
-With the exception of genus `E`, genus representation is balanced in both wild and captive lemurs.
+With the exception of genus `E`, genus (@fig-genus-plots) representation is balanced in both wild and captive lemurs.
 `E` might be high in both cases, as it could be a common genus or have a higher conservation priority.
 
-For birth month, births in the wild are concentrated in a few months, suggesting a breeding season.
+For birth month (@fig-month-born-effects), births in the wild are concentrated in a few months, suggesting a breeding season.
 While in captivity, births are more evenly spread across the year, with some peaks in specific months.
@@
 
-For `Age`, we observe that in the wild, the age distribution of lemurs shows a higher frequency of younger lemurs, with relatively few individuals reaching older ages.
+For `Age` (@fig-age-distribution), we observe that in the wild, the age distribution of lemurs shows a higher frequency of younger lemurs, with relatively few individuals reaching older ages.
 In captivity, the age distribution extends further, with a significant number of lemurs living to older ages.
@@
 # Read data
-wild_data <- read.csv("../data/analysis_data/wild.csv")
-captive_data <- read.csv("../data/analysis_data/captive.csv")
+wild_data <- read_parquet("../data/analysis_data/wild.parquet")
+captive_data <- read_parquet("../data/analysis_data/captive.parquet")
 
@@
 captive_plot_species
-wild_plot_genus + captive_plot_genus + wild_plot_month + captive_plot_month
 ```
 
-We will also look at the distribution of the target variable `Age` in the dataset.
-The following [plot](@fig-age-distribution) shows the distribution of the ages of the lemurs in the dataset for both lemurs in captivity and in the wild.
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-genus-plots
+#| fig-cap: Distribution of Lemurs by Genus
+wild_plot_genus + captive_plot_genus
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-month-plots
+#| fig-cap: Distribution of Lemurs by Birth Month
+
+wild_plot_month + captive_plot_month
+```
 
@@
 
-I used Generalized Linear Model (GLM) framework with a Gaussian family distribution to determine the expected lifespan of a lemur based on the selected variables.
+I used Generalized Linear Model (GLM) framework with a Gaussian family distribution to determine the expected lifespan of a lemur based on the selected variables for both models, for wild and captive.
+`birth type`, captive or wild, was not chosen as a predictor variable because there is significantly less data for wild and we thought it would be easier to visualize the predictions when done seperately.
+However, in doing so, we lost the oppoturnity of studying interaction of `birth_type` with other predictors.
 GLM is a statistical model that generalizes linear regression to include non-normal distributions.
@@
 \end{equation}
-
+<!-- 
 $$
@@
 $$
-where,
+where, -->
 
-- $\bar{a}_\text{wild}$ is the expected lifespan of a lemur in the wild.
-- $\bar{a}_\text{captive}$ is the expected lifespan of a lemur in captivity.
+- $\bar{a}$ is the expected lifespan of a lemur in the wild or captive.
 - $\beta_0$ is the intercept term. It represents the expected lifespan of a lemur when all other variables are zero.
@@
 We chose a Gaussian family distribution for the model because the dependent variable `Age` is continuous and normally distributed.
-From @citenjgov we found that the expected lifespan of a lemur in wild is upto 18 years, and in captivity about 30 years.
+`glm` was used to create the model which is available in the base `R` package.
+The model was then saved using `saveRDS` function to be used for predictions and further analysis.
+<!-- From @citenjgov we found that the expected lifespan of a lemur in wild is upto 18 years, and in captivity about 30 years.
 So, we set the prior for the expected lifespan of a lemur in the wild to be $\mbox{Normal}(16, 4)$ and in captivity to be $\mbox{Normal}(26, 4)$.
 We set the priors for the coefficients to be $\mbox{Normal}(0, 2.5)$, because we do not have any prior information about the effect of the independent variables on the dependent variable.
+-->
+<!-- 
+Using moderately broad priors can help regularize the model, reducing the risk of overfitting and make us have more stable estimates, especially in scenarios with limited data. -->
 
-Using moderately broad priors can help regularize the model, reducing the risk of overfitting and make us have more stable estimates, especially in scenarios with limited data.
+@sec-appendix-model covers coefficients, standard errors, convergence checks and other model diagnostics for our model.
 
-@app_model-dataset covers coefficients, standard errors, convergence checks and other model diagnostics for our model.
+A combined model was tried, where `birth_type` was taken as the predictor variable, but it had worse residual and p values, so we decided to go with two separate models for wild and captive lemurs.
+The seperate models allow us to better understand the factors affecting the lifespan of lemurs in each environment, making them easy to visualize.
+At the expense of, not being able to study the interaction of `birth_type` with other predictors.
 
@@
 #| label: fig-sex-effects
-#| fig-cap: Comparison of Average Predicted Ages of Lemurs by Age Range, Sex, and Prediction Type (Wild vs. Captive)
-
-# Define age ranges
-predictions <- predictions %>%
-  mutate(age_range = cut(age, breaks = c(0, 10, 20, 40), labels = c("0-10", "10-20", "20+"), include.lowest = TRUE))
+#| fig-cap: Comparison of Average Predicted Ages of Lemurs by Sex of Wild and Captive Lemurs
 
@@
 plot_data <- predictions %>%
-  group_by(age_range, sex) %>%
+  group_by(sex) %>%
   summarise(
@@
 
-# remove NAs
+# Remove NAs
 plot_data <- plot_data %>% drop_na()
@@
 # Create the bar plot
-plotted <- ggplot(plot_data, aes(x = age_range, y = avg_age, fill = prediction_type)) +
+plotted <- ggplot(plot_data, aes(x = sex, y = avg_age, fill = prediction_type)) +
   geom_bar(stat = "identity", position = "dodge") +
-  facet_wrap(~sex) +
   scale_fill_manual(
@@
   labs(
-    x = "Age Ranges",
+    x = "Sex",
     y = "Average Predicted Age",
@@
 
-# Discussion {#sec-discussion}
+# Discussion
 
-In this study, we sought to predict and analyze the lifespan of lemurs in wild and captive environments by examining factors of sex, species, genus, and birth month.
-Utilizing data from the Duke Lemur Center, we divided the dataset into wild-born and captive-born lemurs to understand how environmental conditions impact longevity.
-Generalized Linear Models (GLMs) with a Gaussian distribution were employed to understand the relationships between these variables and lifespan.
+This paper investigates the lifespan of lemurs across wild and captive environments, utilizing data from the Duke Lemur Center (DLC).
+Generalized Linear Models (GLMs) with Gaussian distribution are used to analyze how factors such as sex, species, genus, and birth month impact lemur longevity.
+Separate models are developed for wild-born and captive-born lemurs to account for environmental differences.
+The study identifies patterns of increased longevity in captive lemurs and emphasizes species-specific conservation strategies.
 
 ## Key Findings 
+One key insight is that captivity significantly increases lemur lifespan.
+Controlled environments provide consistent access to food, medical care, and protection from predators, reducing environmental stresses that contribute to shorter lifespans in the wild.
+This finding underscores the importance of captivity as a conservation tool for endangered species like lemurs.
 
-We learn from this study that captivity plays a major role in the lifespan of lemurs.
-Captive lemurs, benefit from controlled environments, consistent food supplies, and medical care, live considerably longer on average than their wild counterparts similar to as the study by @citetidi showed.
-This underscores the potential of captive breeding programs to contribute positively to species survival and highlights the role of human intervention in extending the lifespans of endangered species.
+Species-specific differences in lifespan emerge as a critical factor.
+Certain species thrive in captivity due to tailored conservation practices, while others exhibit shorter lifespans, suggesting room for improvement.
+Additionally, the diminished influence of sex on lifespan in captivity contrasts with its pronounced effects in the wild, revealing the role of environmental factors in mediating biological differences.
 
-Another important finding is the variability in lifespan based on species and sex.
-In the wild, male lemurs tend to live longer than females, possibly due to biological or behavioral factors.
-However, this sex-based difference diminishes in captivity, suggesting that controlled environments mitigate sex-specific survival challenges.
-Additionally, species differences in longevity were significant in both settings, highlighting the influence of inherent biological factors and emphasizing the need for species-specific conservation strategies.
-
-## Weaknesses and Limitations
-
-Despite these findings, this study has several limitations.
-The dataset used was relatively small for wild-born lemurs, potentially affecting the reliability of the statistical models.
-For instance, the limited sample size may have led to the inclusion of predictors with p-values slightly above the conventional threshold, which could inflate Type I error rates.
-Furthermore, the data on wild-born lemurs lacked detailed information on maternal and paternal species and litter size, which are potentially critical factors in understanding lifespan dynamics, and we could not include them in our analysis.
-
-Another limitation lies in the potential bias introduced by data collection processes.
-The reliance on historical records and the transition between database systems may have resulted in inconsistencies or inaccuracies.
-Additionally, the study only considered certain categorical variables, excluding other possible factors such as habitat quality, diet variability, or exposure to diseases, which could provide a more holistic understanding of lemur longevity.
+## Drawbacks of the Study
+While the study provides valuable insights, it has limitations.
+First, the dataset is skewed towards captive lemurs, limiting the robustness of wild lifespan analysis.
+Second, the exclusion of interactions between birth type and other predictors in the combined model reduces the depth of insights.
+Lastly, some predictors, like birth month, have inconsistent or weak effects, indicating potential noise in the data or the need for more refined models.
 
 ## Future Directions
+Future research should focus on:
+1. Expanding datasets to include more wild lemurs for balanced comparisons.
+2. Exploring interactions between predictors (e.g., birth type and species) to uncover complex relationships.
+3. Investigating why some species perform better in the wild than in captivity, which could inform more effective conservation strategies.
+4. Examining seasonal effects and birth month impacts further to identify environmental triggers influencing lifespan.
+5. Using advanced modeling techniques like Bayesian models or neural networks for more nuanced predictions.
 
-This research opens the door to several avenues for further investigation.
-Future studies should aim to collect larger and more representative datasets for wild-born lemurs to improve the robustness of the models.
-Additionally, incorporating environmental variables such as habitat degradation, climate patterns, and predation risks could provide better understanding of the challenges faced by wild populations.
-
-Furthermore, understanding why certain species thrive better in the wild (e.g., ALB, COR, MAC, and MON) compared to captivity requires focused investigation.
-This could involve studying behavioral adaptations, ecological niches, and species-specific physiological traits.
-Expanding the analysis to include genetic data may also uncover underlying biological factors that contribute to these differences.
-
-Finally, longitudinal studies examining the long-term impacts of captivity on lemur behavior, reproduction, and overall health would be invaluable.
-Integrating interdisciplinary approaches, combining ecological, genetic, and conservation sciences, could help develop more effective strategies for preserving these critically endangered species.
-
-In conclusion, while this study provides meaningful insights into the factors influencing lemur lifespan, it also highlights the complexities of their survival in diverse environments.
-Addressing the limitations and expanding the scope of research will not only deepen our understanding but also support more informed conservation efforts for lemurs and other endangered species.
+By addressing these areas, conservationists can develop targeted strategies to enhance the survival of lemurs both in the wild and in captivity.
 
@@
 
-wild_data <- read.csv("../data/analysis_data/wild.csv")
-captive_data <- read.csv("../data/analysis_data/captive.csv")
+wild_data <- read_parquet("../data/analysis_data/wild.parquet")
+captive_data <- read_parquet("../data/analysis_data/captive.parquet")
 
@@
 
+# Model Summary {#sec-appendix-model}
+
+## Wild Lemurs' Model
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-wild-model-coefficients
+#| tbl-cap: Summary of the model for wild lemurs
+
+# Extract coefficients
+coefficients <- broom::tidy(wild_model, conf.int = TRUE)
+coefficients |> kable(digits = 2)
+```
+
+## Captive Lemurs' Model
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-captive-model-coefficients
+#| tbl-cap: Summary of the model for captive lemurs
+
+# Extract coefficients
+coefficients <- broom::tidy(captive_model, conf.int = TRUE)
+coefficients |> kable(digits = 2)
+```
+
+# Posterior Predictive Checks {#sec-appendix-ppc}
+
+## Residual Plots
+
+As show in @fig-residual-plots, the residual plots for both captive and wild models show no clear patterns, indicating that the models are capturing the underlying relationships between the predictors and the response variable effectively.
+The residuals are randomly distributed around zero, with no discernible trends or heteroscedasticity, suggesting that the model assumptions are met.
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-residual-plots
+#| fig-cap: Residual plots for the wild and captive models
+# Create residual plot for wild model
+wild_plot <- ggplot(data = data.frame(Fitted = wild_model$fitted.values, 
+                                      Residuals = residuals(wild_model)),
+                    aes(x = Fitted, y = Residuals)) +
+  geom_point() +
+  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
+  labs(title = "Residual Plot for Wild Model", x = "Fitted Values", y = "Residuals") +
+  theme_minimal()
+
+# Create residual plot for captive model
+captive_plot <- ggplot(data = data.frame(Fitted = captive_model$fitted.values, 
+                                         Residuals = residuals(captive_model)),
+                       aes(x = Fitted, y = Residuals)) +
+  geom_point() +
+  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
+  labs(title = "Residual Plot for Captive Model", x = "Fitted Values", y = "Residuals") +
+  theme_minimal()
+
+wild_plot + captive_plot
+
+```
+
+## QQ Plots
+As show in @fig-qq-plots, the QQ plots for both captive and wild models show that the standardized residuals closely follow the expected normal distribution line.
+The points are mostly aligned along the line, indicating that the residuals are normally distributed.
+This suggests that the model assumptions of normality are met, and the models are appropriate for the data.
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-qq-plots
+#| fig-cap: QQ plots for the wild and captive models
+
+# Create QQ plot for wild model
+wild_qq <- ggplot(data = data.frame(Standardized_Residuals = rstandard(wild_model)),
+                  aes(sample = Standardized_Residuals)) +
+  stat_qq() +
+  stat_qq_line() +
+  labs(title = "QQ Plot for Wild Model") +
+  theme_minimal()
+
+# Create QQ plot for captive model
+captive_qq <- ggplot(data = data.frame(Standardized_Residuals = rstandard(captive_model)),
+                     aes(sample = Standardized_Residuals)) +
+  stat_qq() +
+  stat_qq_line() +
+  labs(title = "QQ Plot for Captive Model") +
+  theme_minimal()
+
+wild_qq + captive_qq
+```
 #  Species of Lemurs {#sec-appendix-species}
@@
 \newpage
+
+# References {#sec-references}

### 2024-11-28 scripts/02-download_data.r

@@ -17,3 +17,4 @@ if (file.exists(destination_file)) {
 } else {
   cat("Failed to download the file.\n")
 }
+

### 2024-11-28 scripts/03-clean_data.r

@@ -57,4 +57,4 @@ write_parquet(data_wild, "data/analysis_data/wild.parquet")
 write_parquet(data_captive, "data/analysis_data/captive.parquet")
 
 
-unique(data$genus)
+unique(data$species)

### 2024-11-28 scripts/05-data-analysis.r

@@ -8,9 +8,7 @@
 # Pre-requisites: None
 
 ## Workspace setup ##
-library(tidyverse)
 library(arrow)
-library(rstanarm)
 
 # Load data
 wild_data <- read_parquet("data/analysis_data/wild.parquet")
@@ -22,6 +20,13 @@ wild_model <- glm(
     family = gaussian(),
     data = wild_data,
 )
+
+my_model <- glm(
+    formula = age ~ species + sex + birth_type + month_born + genus,
+    family = gaussian(),
+    data = data,
+)
+
 captive_model <- glm(
     formula = age ~ species + sex + month_born + genus,
     family = gaussian(),
@@ -31,3 +36,5 @@ captive_model <- glm(
 # save models
 saveRDS(wild_model, "models/wild_model.rds")
 saveRDS(captive_model, "models/captive_model.rds")
+
+

### 2024-11-28 scripts/06-predicting.r

@@ -18,14 +18,12 @@ wild_model <- readRDS("models/wild_model.rds")
 captive_model <- readRDS("models/captive_model.rds")
 
 # creating prediction dataset
-ages_ <- unique(analysis_data$age)
 species_ <- unique(analysis_data$species)
 sex_ <- unique(analysis_data$sex)
 month_born_ <- unique(analysis_data$month_born)
 genus_ <- unique(analysis_data$genus)
 
 prediction_data <- expand.grid(
-    age = ages_,
     species = species_,
     sex = sex_,
     month_born = month_born_,
@@ -38,7 +36,6 @@ predictions_captive <- predict(captive_model, newdata = prediction_data, type = 
 
 
 predictions_df <- data.frame(
-    age = prediction_data$age,
     species = prediction_data$species,
     sex = prediction_data$sex,
     month_born = prediction_data$month_born,
@@ -49,3 +46,4 @@ predictions_df <- data.frame(
 
 # save the predictions
 write_csv(predictions_df, "data/predictions/predictions.csv")
+

### 2024-11-28 scripts/figure.r

@@ -1,41 +1,24 @@
-library(ggplot2)
-library(dplyr)
+library(arrow)
+# load model
+wild_model <- readRDS("models/wild_model.rds")
+wild_data <- read_parquet("data/analysis_data/wild.parquet")
+captive_model <- readRDS("models/captive_model.rds")
 
-WILD_COLOR <- "#FFA500"
-CAPTIVE_COLOR <- "#4169E1"
+# Residual plots for wild_model
+plot(wild_model$fitted.values, residuals(wild_model), main="Residual Plot for Wild Model")
+abline(h=0, col="red")
 
-# Load the data
-predictions <- read_csv("data/predictions/predictions.csv")
-# Summarize the data by month
-month_plot_data <- predictions %>%
-  group_by(month_born) %>%
-  summarise(
-    avg_age_wild = mean(predicted_age_wild, na.rm = TRUE),
-    avg_age_captive = mean(predicted_age_captive, na.rm = TRUE)
-  ) %>%
-  pivot_longer(
-    cols = c(avg_age_wild, avg_age_captive),
-    names_to = "prediction_type",
-    values_to = "avg_age"
-  )
+# QQ plot to check normality of residuals
+qqnorm(residuals(wild_model))
+qqline(residuals(wild_model), col="blue")
 
-# Create the bar plot
-month_plot <- ggplot(month_plot_data, aes(x = month_born, y = avg_age, fill = prediction_type)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  scale_fill_manual(
-    values = c("avg_age_wild" = WILD_COLOR, "avg_age_captive" = CAPTIVE_COLOR),
-    labels = c("Wild Lemurs", "Captive Lemurs")
-  ) +
-  labs(
-    x = "Month Born",
-    y = "Average Predicted Age",
-    fill = "Type of Lemurs",
-    title = "Average Predicted Ages by Month Born"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# Residual plots for captive_model
+wild_model$aic
 
-month_plot
+install.packages("car")
+library(car)
+vif(wild_model)
 
-# save the plot
-ggsave("figures/month_plot.png", month_plot, width = 10, height = 6, units = "in", dpi = 300)
+wild_predictions <- predict(wild_model, type="response")
+plot(wild_data$age, wild_predictions, main="Predicted vs Observed (Wild Model)")
+abline(0, 1, col="red")

### 2024-12-04 paper/paper.qmd
Grammar correcitons in the paper. Figure captions added

@@
 date-format: long
-abstract: "This study investigates the lifespan of lemurs in both wild and captive environments using data from the Duke Lemur Center, aiming to identify how factors such as sex, species, genus, and birth month influence longevity.
+abstract: "This study investigates the lifespan of lemurs in wild and captive environments using data from the Duke Lemur Center, aiming to identify how factors such as sex, species, genus, and birth month influence longevity.
 Using Generalized Linear Models, the analysis highlights that lemurs in captivity live significantly longer than their wild counterparts, emphasizing the impact of conservation practices.
@@
 library(rstanarm)
+library(knitr)
+
 # Set the color for lemurs
@@
 
+# Read data
+wild_data <- read_parquet("../data/analysis_data/wild.parquet")
+captive_data <- read_parquet("../data/analysis_data/captive.parquet")
+# Load predictions
 predictions <- read_csv("../data/predictions/predictions.csv")
+# Load models
 wild_model <- readRDS("../models/wild_model.rds")
@@
 Understanding the factors that influence their lifespan is crucial for guiding effective conservation efforts.
-The Duke Lemur Center (DLC), established in 1966, has played a pivotal role in the study and preservation of lemurs, maintaining the largest population of these species outside their native Madagascar.
-Different factors can affect the longevity of lemurs in wild and captive, longevity is important because the longer lemurs live, the more likely they are to reproduce and contribute to the survival of their species.
+The Duke Lemur Center (DLC), established in 1966, has played a pivotal role in the study and preservation of lemurs, maintaining the largest population of these species outside their native region Madagascar.
+Different factors can affect the longevity of lemurs in wild and captive.
+Longevity is important because the longer lemurs live, the more likely they are to reproduce and hence, contribute to the survival of their species.
 
-This study uses a dataset from the DLC to analyze if and how sex, species, genus and month a lemur is born in can affect its lifespan in the wild and captivity.
-This study can help us guide what kinds of lemur are we failing to conserve in captivity and how they perform in wild.
-It will also reinforce where our efforts have been successful.
+My study uses a dataset from the DLC to analyze if and how sex, species, genus and month a lemur is born in can affect its lifespan in the wild and captivity.
+This study can provide insights into which lemur species we are struggling to conserve in captivity and their performance in the wild.
+It will highlight areas where our conservation efforts have been effective.
 
@@
 
-The remainder of this paper is structured as follows: @sec-data details the data collection and processing methods, while @sec-model outlines the modeling approach.
+The remainder of this paper is structured as follows: @sec-data details the data collection and processing methods.
+@sec-model outlines the modeling approach.
 @sec-results presents the results, highlighting key differences between wild and captive lemurs.
-The discussion in @sec-sec-discussion contextualizes these findings, addressing limitations and proposing future directions for research.
-Finally, the @sec-appendixs includes additional information on data cleaning, the analysis dataset, and the model dataset.
+The discussion in @sec-discussion contextualizes these findings, addressing limitations and proposing future directions for research.
+Finally, @sec-appendix is the appendix and includes additional information on data cleaning, the analysis dataset, and the model dataset.
 
 # Data {#sec-data}
-The data was taken from @cookson_duke_lemur_center, who acquired it from Duke Lemur Center more about which, can be found in @sec-data_collection.
+The data was taken from the GitHub repository of @cookson_duke_lemur_center, who acquired it from Duke Lemur Center.
+Details on how data was acquired and processed can be found in @sec-data_collection.
 We used `R` [@citeR] along with its packages to clean and analyze the data.
-These packages include `tidyverse` [@cite-tidyverse], `ggplot2` [@cite-ggplot], `dplyr` [@cite-dplyr], `patchwork` [@cite-patchwork], `arrow` [@cite-arrow], `knitr` [@cite-knitr] and `rstanarm` [@cite-rstanarm].
+These packages include `tidyverse` [@cite-tidyverse], `ggplot2` [@cite-ggplot], `dplyr` [@cite-dplyr], `patchwork` [@cite-patchwork], `arrow` [@cite-arrow], `knitr` [@cite-knitr], `rstanarm` [@cite-rstanarm] and `kableExtra` [@cite-kableExtra].
 
@@
 
-- `Sex`: The sex of the lemur on birth. It is a cateogrical variable that can be `M` or `F`.
+- `Sex`: The sex of the lemur on birth. It is a cateogrical variable that can be `M` or `F`. I took sex as a factor because it is common to see females tend to live longer [@cite-Lemaitre2020].
 - `Species`: The species of the lemur.
 It is also a categorical variable and can take one of the following values: `GG`, `COL`, `UL`, `RUF`, `MOH`, `MAC`, `CAT`, `FUL`, `ALB`, `AR`, `VV`, `COQ`, `MED`, `MUR`, `COU`, `TAR`, `PYG`, `ZAZ`, `MON`, `RUB`, `COR`, `SAN`, `FLA`, `MAD`, `POT`.
-The specific or common names of these species can be found in Appendix [A](#appendix_a).
-- `Genus`: The genus of the lemur. It is a categorical variable and can take one of the following values: `O`, `E`, `G`, `H`, `L`, `V`, `P`, `C`, `M`, `N`, `D`.
+The specific or common names of these species can be found in @sec-appendix-species.
+Species is important as different species have different requirements and threats, which can affect their lifespan.
+- `Genus`: The genus of the lemur.
+It is a categorical variable and can take one of the following values: `O`, `E`, `G`, `H`, `L`, `V`, `P`, `C`, `M`, `N`, `D`.
+Genus can have the same effect as species.
 - `month_born`: The month in which the lemur was born.
 It is a categorical variable and can take one of the following values: `1`, `2`, `3`, `4`, `5`, `6`, `7`, `8`, `9`, `10`, `11`, `12`, where `1` corresponds to January, `2` to February, and so on.
+Month born can be important as it can affect the availability of food and other resources during its early stages, which can affect the lifespan of the lemur.
 
@@
 
-The following figures show the distribution of the lemurs in the dataset based on the variables `Sex` (@fig-sex-plots), `Species` (@fig-species-effects), `Genus` (@fig-genus-plots), and `month_born` (@fig-month-born-effects) for both lemurs in captivity, shown in orange and in the wild, shown in dark green.
+The following figures show the distribution of the lemurs in the dataset based on the variables `Sex` (@fig-sex-plots), `Species` (@fig-species-plots and @fig-species-captive-plots), `Genus` (@fig-genus-plots), and `month_born` (@fig-month-plots) for both lemurs in captivity, shown in orange and in the wild, shown in dark green.
 
-For `Sex` (@fig-sex-plots), we observe that in the wild, the distribution of male and female lemurs is nearly even, with only a slight skew.
-While in captivity, there are significantly more female lemurs than males.
-The larger count of female lemurs in captivity could result from conservation breeding programs prioritizing females to ensure the species' survival.
+For `Sex` (@fig-sex-plots), we observe that in the wild, there are more females than males.
+This might be due to the fact that females tend to live longer than males for most mammals [@cite-Lemaitre2020].
+While in captivity, there are significantly more male lemurs than males.
+This shows a short coming in the conservation efforts, as we are not able to conserve the females, which are more important for the survival of the species.
 
@@
 In contrast, the challenges of the wild, such as predation, disease, and fluctuating resources, contribute to shorter lifespans.
-With there being one exception, we observe signigicant number of captive lemurs below age 1, this can be due to the extensive breeding of captivity or could be a sign of data quality issues.
 
@@
 #| label: fig-count-plots
-
-# Read data
-wild_data <- read_parquet("../data/analysis_data/wild.parquet")
-captive_data <- read_parquet("../data/analysis_data/captive.parquet")
 
@@
 #| label: fig-sex-plots
-#| fig-cap: Distribution of Lemurs by Sex
+#| fig-cap: Sex distribution of wild and captive lemurs represented by bar plots for female (F) and male (M) individuals.
 # Display plots
@@
 #| label: fig-species-plots
+#| fig-cap: Species-wise count of wild lemurs in our dataset.
 wild_plot_species
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-species-captive-plots
+#| fig-cap: Species-wise count of captive lemurs in our dataset.
+
 captive_plot_species
@@
 #| label: fig-genus-plots
-#| fig-cap: Distribution of Lemurs by Genus
+#| fig-cap: Distribution of lemurs by genus in the wild (on left) and in captivity (on right).
 wild_plot_genus + captive_plot_genus
@@
 #| label: fig-month-plots
-#| fig-cap: Distribution of Lemurs by Birth Month
-
+#| fig-cap: Distribution of lemurs by the month they were born in. On left is the distribution for lemurs in the wild and on the right is for lemurs in captivity.
 wild_plot_month + captive_plot_month
@@
 #| label: fig-age-distribution
-#| fig-cap: Distribution of ages of lemurs in the dataset for both lemurs in captivity and in the wild.
+#| fig-cap: Distribution of ages of lemurs in the dataset for lemurs in captivity (on right) and in the wild (on left).
 
@@
   geom_histogram(fill = WILD_COLOR, bins = 30) +
-  labs(title = "Wild Lemurs",
+  labs(
        x = "Age",
@@
   geom_histogram(fill = CAPTIVE_COLOR, bins = 30) +
-  labs(title = "Captive Lemurs",
+  labs(
        x = "Age",
@@
 The DLC created two data files from the database: **the DLC Animal List**, which contains single-copy variables for each animal in the colony's history, and **the DLC Weight File**, which contains all weight measurements for each animal.
-We used the DLC Animal List for this analysis, which was first cleaned by @cookson_duke_lemur_center and then we clean it as per our needs.
+We used the DLC Animal List for this analysis, which was first cleaned by @cookson_duke_lemur_center and then I clean it.
+The steps I took for cleaning the data can be found in @sec-appendix-data-cleaning.
 
@@
 
-I used Generalized Linear Model (GLM) framework with a Gaussian family distribution to determine the expected lifespan of a lemur based on the selected variables for both models, for wild and captive.
-`birth type`, captive or wild, was not chosen as a predictor variable because there is significantly less data for wild and we thought it would be easier to visualize the predictions when done seperately.
-However, in doing so, we lost the oppoturnity of studying interaction of `birth_type` with other predictors.
+I used Generalized Linear Model (GLM) framework with a Gaussian family distribution to determine the expected lifespan of a lemur based on the selected variables for both models, wild and captive.
+`birth type` (captive or wild) was not chosen as a predictor variable because there is significantly less data for wild and I thought it would be easier to visualize the predictions when done seperately.
+However, in doing so, we lost the oppoturnity of studying interaction effects of `birth_type` with other predictors.
+
 GLM is a statistical model that generalizes linear regression to include non-normal distributions.
@@
 
-My model will be based on `Sex`, `Species`, `Genus`, and `month_born` as the independent variables and `Age` as the dependent variable.
+My model is based on `Sex`, `Species`, `Genus`, and `month_born` as the independent variables and `Age` as the dependent variable.
 
@@
 \end{equation}
-<!-- 
-$$
-\begin{aligned}
-\bar{a}_\text{wild} & \sim \mbox{Normal}(16, 4)\\
-\bar{a}_\text{captive} & \sim \mbox{Normal}(26, 4)\\
-\beta_0 & \sim \mbox{Normal}(0, 2.5)\\
-\beta_1 & \sim \mbox{Normal}(0, 2.5)\\
-\beta_2 & \sim \mbox{Normal}(0, 2.5)\\
-\beta_3 & \sim \mbox{Normal}(0, 2.5)\\
-\beta_4 & \sim \mbox{Normal}(0, 2.5)
-\end{aligned}
-$$
-where, -->
+
+where:
 
 - $\bar{a}$ is the expected lifespan of a lemur in the wild or captive.
+
 - $\beta_0$ is the intercept term. It represents the expected lifespan of a lemur when all other variables are zero.
+
 - $\beta_1$ is the coefficient for `Sex`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Sex`.
+
 - $\beta_2$ is the coefficient for `Species`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Species`.
+
 - $\beta_3$ is the coefficient for `Genus`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Genus`.
+
 - $\beta_4$ is the coefficient for `month_born`. It represents the change in the expected lifespan of a lemur for a one-unit change in `month_born`.
+
 - $\epsilon$ is the error term.
 
-We chose a Gaussian family distribution for the model because the dependent variable `Age` is continuous and normally distributed.
+I chose a Gaussian family distribution for the model because the dependent variable `Age` is continuous and normally distributed.
 `glm` was used to create the model which is available in the base `R` package.
 The model was then saved using `saveRDS` function to be used for predictions and further analysis.
-<!-- From @citenjgov we found that the expected lifespan of a lemur in wild is upto 18 years, and in captivity about 30 years.
-So, we set the prior for the expected lifespan of a lemur in the wild to be $\mbox{Normal}(16, 4)$ and in captivity to be $\mbox{Normal}(26, 4)$.
-We set the priors for the coefficients to be $\mbox{Normal}(0, 2.5)$, because we do not have any prior information about the effect of the independent variables on the dependent variable.
--->
-<!-- 
-Using moderately broad priors can help regularize the model, reducing the risk of overfitting and make us have more stable estimates, especially in scenarios with limited data. -->
 
-@sec-appendix-model covers coefficients, standard errors, convergence checks and other model diagnostics for our model.
+@sec-appendix-model covers coefficients, standard errors, convergence checks and other model diagnostics of our model.
 
-A combined model was tried, where `birth_type` was taken as the predictor variable, but it had worse residual and p values, so we decided to go with two separate models for wild and captive lemurs.
+A combined model was tried, where `birth_type` was taken as the predictor variable, but it had worse residual and p values, so I decided to go with two separate models for wild and captive lemurs.
 The seperate models allow us to better understand the factors affecting the lifespan of lemurs in each environment, making them easy to visualize.
-At the expense of, not being able to study the interaction of `birth_type` with other predictors.
+At the expense of, not being able to study the interaction effects of `birth_type` with other predictors.
 
@@
 #| label: fig-wild-model-coefficients
-#| fig-cap: Coefficients of the wild model
-
-wild_model <- readRDS("../models/wild_model.rds")
+#| fig-cap: Significant coefficients (p-value < 0.07) with their 95% credible intervals for the wild model.
 
@@
 #| label: fig-captive-model-coefficients
-#| fig-cap: Coefficients of the captive model
-
-captive_model <- readRDS("../models/captive_model.rds")
+#| fig-cap: Significant coefficients (p-value < 0.05) with their 95% credible intervals for the captive model.
 
@@
 Few species in the wild have significant effect on the age, these include `VV`, `SAN`, `RUF`, `RUB`, `MAD` and `CAT`.
-For sex, as expected in wild, males have a higher lifespan than females.
-We had to jump over p-value of 0.07 to get the significant results, as we had limited data in case of wild.
+`SAN`, `RUF` and `RUB` have a negative effect on age, while `VV`, `MAD` and `CAT` have a positive effect.
+
+For sex, in wild, males have a higher lifespan than females.
+Higher lifespan of males suggest that the more number of females observed in the wild in @fig-sex-plots might have other underlying cause than just females living longer, which can be a good area of research for Duke Lemur Center.
+
+Note that we had to increase our threshold of p-value to 0.07 to get the significant results, as we had limited data in case of wild.
 
@@
 Few species have a negative effect on age, these include `MUR`, `MOH`, `MAD`, `GG`, `COQ` and `AR`.
-In case of captivity, sex is not a major factor in age, as its confidence interval is nearly centered at 0.
-Overall, species differences are prominent, while birth month effects appear less consistent, mostly centered around 0.
-Except December and September seem to be negatively affecting the age of lemurs in captivity, while November positively.
+I did not find any specific species with positive effect on age in captivity.
+This is a good sign as it shows that the conservation efforts are working for most of the species in captivity.
+
+In case of captivity, sex is not a major factor in age, as its confidence interval is nearly centered at 0.  Another good sign towards conservation effects.
+
+Birth month effects appear less consistent, mostly centered around 0.
+Except December and September that seem to be negatively affecting the age of lemurs in captivity, while November slightly positively.
 
@@
 #| label: fig-sex-effects
-#| fig-cap: Comparison of Average Predicted Ages of Lemurs by Sex of Wild and Captive Lemurs
+#| fig-cap: Comparison of average predicted ages of wild and captive lemurs across sexes. Males live longer in both wild and captive environments, but the difference is more pronounced in captivity.
 
@@
 
-From the @fig-sex-effects, lemurs in captive have a higher average predicted age than those in the wild across all age ranges.
+From the @fig-sex-effects, lemurs in captive have a higher average predicted age than those in the wild across sexes.
 And in the captive, we see males living longer on average than females, while in the wild, the difference is less pronounced.
@@
 #| label: fig-month-born-effects
-#| fig-cap: Comparison of Average Predicted Ages of Lemurs by Age Range, Birth Month, and Prediction Type (Wild vs. Captive)
+#| fig-cap: Comparison of average predicted ages of wild and captive lemurs across different months born.
 
@@
     fill = "Type of Lemurs",
-    title = "Average Predicted Ages by Month Born"
   ) +
@@
 #| label: fig-species-effects
-#| fig-cap: Comparison of Average Predicted Ages of Lemurs by Age Range, Species, and Prediction Type (Wild vs. Captive)
+#| fig-cap: Comparison of average predicted ages of wild and captive lemurs across different species.
+Captive lemurs generally show a higher predicted age than their wild counterparts for most species.
 
@@
     fill = "Type of Lemurs",
-    title = "Average Predicted Ages by Species"
   ) +
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 
@@
 Species-specific differences in lifespan emerge as a critical factor.
-Certain species thrive in captivity due to tailored conservation practices, while others exhibit shorter lifespans, suggesting room for improvement.
-Additionally, the diminished influence of sex on lifespan in captivity contrasts with its pronounced effects in the wild, revealing the role of environmental factors in mediating biological differences.
+Most of the species thrive in captivity due to tailored conservation practices, while some including `ALB`, `COR`, `MAC`, and `MON` perform better in the wild.
+Understanding the reasons behind these variations can inform targeted conservation efforts to improve outcomes.
+
+Even though no specific trend was observed for birth month, lemurs born in August and November tend to live longer in the wild than in captivity. It is worth exploring the underlying cause for this.
+
+Additionally, the greater influence of sex on lifespan in captivity contrasts with its less pronounced effects in the wild opens up avenues for further research and mitigating any short comings in our conservation efforts.
 
 ## Drawbacks of the Study
-While the study provides valuable insights, it has limitations.
-First, the dataset is skewed towards captive lemurs, limiting the robustness of wild lifespan analysis.
+While the study provides valuable insights, it has its limitations. First, the dataset is skewed towards captive lemurs, limiting the robustness of wild lifespan analysis.
+
 Second, the exclusion of interactions between birth type and other predictors in the combined model reduces the depth of insights.
-Lastly, some predictors, like birth month, have inconsistent or weak effects, indicating potential noise in the data or the need for more refined models.
+
+Third, some predictors, like birth month, have inconsistent or weak effects, indicating potential noise in the data or the need for more refined models.
+
+Finally, the ages of lemurs below one year were excluded, which may have impacted the analysis of early-life mortality and its effects on lifespan.
 
@@
 Future research should focus on:
+
 1. Expanding datasets to include more wild lemurs for balanced comparisons.
+
 2. Exploring interactions between predictors (e.g., birth type and species) to uncover complex relationships.
+
 3. Investigating why some species perform better in the wild than in captivity, which could inform more effective conservation strategies.
+
 4. Examining seasonal effects and birth month impacts further to identify environmental triggers influencing lifespan.
-5. Using advanced modeling techniques like Bayesian models or neural networks for more nuanced predictions.
+
+5. Studying the infant mortality rate and its impact on overall lemur lifespan.
+
+6. Using advanced modeling techniques like Bayesian models or neural networks for more nuanced predictions.
 
 By addressing these areas, conservationists can develop targeted strategies to enhance the survival of lemurs both in the wild and in captivity.
+Such studies help us understand what are we doing right in conservation efforts, and where should we focus next.
 
@@
 #| label: tbl-wild
-#| tbl-cap: Analysis dataset for Wild Lemurs
-library(knitr)
-
-wild_data <- read_parquet("../data/analysis_data/wild.parquet")
-captive_data <- read_parquet("../data/analysis_data/captive.parquet")
+#| tbl-cap: Analysis dataset for wild lemurs
 
@@
 #| label: tbl-captive
-#| tbl-cap: Analysis dataset for Captive Lemurs
+#| tbl-cap: Analysis dataset for captive lemurs
 
@@
 #| label: tbl-captive-model-coefficients
+
 #| tbl-cap: Summary of the model for captive lemurs
@@
 #| label: fig-residual-plots
-#| fig-cap: Residual plots for the wild and captive models
+#| fig-cap: Residual plots for the wild (on left) and captive (on right) models.
+Both the plots indicate that the residuals are randomly distributed around zero, with no discernible patterns or heteroscedasticity.
 # Create residual plot for wild model
@@
   geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
-  labs(title = "Residual Plot for Wild Model", x = "Fitted Values", y = "Residuals") +
+  labs(x = "Fitted Values", y = "Residuals") +
   theme_minimal()
@@
   geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
-  labs(title = "Residual Plot for Captive Model", x = "Fitted Values", y = "Residuals") +
+  labs(x = "Fitted Values", y = "Residuals") +
   theme_minimal()
@@
 As show in @fig-qq-plots, the QQ plots for both captive and wild models show that the standardized residuals closely follow the expected normal distribution line.
-The points are mostly aligned along the line, indicating that the residuals are normally distributed.
+The QQ plot for wild lemurs show slight deviations at the tails, but overall, the points are aligned along the line.
+Overall, the points are mostly aligned along the line, indicating that the residuals are normally distributed.
 This suggests that the model assumptions of normality are met, and the models are appropriate for the data.
@@
 #| label: fig-qq-plots
-#| fig-cap: QQ plots for the wild and captive models
+#| fig-cap: QQ plots of standardized residuals vs the expected normal distribution for the wild (on left) and captive (on right) models.
+The points closely follow the expected normal distribution line, indicating that the residuals are normally distributed.
 
@@
   stat_qq_line() +
-  labs(title = "QQ Plot for Wild Model") +
   theme_minimal()
@@
   stat_qq_line() +
-  labs(title = "QQ Plot for Captive Model") +
   theme_minimal()
@@
 ```
+
 #  Species of Lemurs {#sec-appendix-species}

### 2024-12-04 scripts/figure.r
readme corrected, unecessary script removed

@@ -1,24 +1,0 @@
-library(arrow)
-# load model
-wild_model <- readRDS("models/wild_model.rds")
-wild_data <- read_parquet("data/analysis_data/wild.parquet")
-captive_model <- readRDS("models/captive_model.rds")
-
-# Residual plots for wild_model
-plot(wild_model$fitted.values, residuals(wild_model), main="Residual Plot for Wild Model")
-abline(h=0, col="red")
-
-# QQ plot to check normality of residuals
-qqnorm(residuals(wild_model))
-qqline(residuals(wild_model), col="blue")
-
-# Residual plots for captive_model
-wild_model$aic
-
-install.packages("car")
-library(car)
-vif(wild_model)
-
-wild_predictions <- predict(wild_model, type="response")
-plot(wild_data$age, wild_predictions, main="Predicted vs Observed (Wild Model)")
-abline(0, 1, col="red")

### 2024-12-04 paper/paper.qmd
title and llm suage updated

@@
 ---
-title: "Predicting a Lemur's Lifespan in Wild and Captivity"
-# subtitle: "My subtitle if needed"
+title: "Lifespan Analysis of Lemurs in Wild and Captive Environments"
+subtitle: "Evaluating Effective Strategies and Highlighting Areas for Conservation Focus"
 author: 
@@
 
+\appendix
+
 # Appendix {#sec-appendix}
-
-\appendix
 

### 2024-12-04 paper/paper.qmd
cleaning scripts, survey appendix added in paper

@@
 Using Generalized Linear Models, the analysis highlights that lemurs in captivity live significantly longer than their wild counterparts, emphasizing the impact of conservation practices.
-The findings reveal species-specific variations in lifespan and the diminishing influence of sex differences in captivity.
+The findings show species-specific variations in lifespan and the pronounced influence of sex differences in captivity.
 These insights help identify effective practices in captivity and suggest areas for improvement to optimize conservation strategies, contributing to the survival of this endangered mammalian group."
@@
 
-Using Generalized Linear Models (GLMs) with Gaussian distribution, factors including sex, species, genus, and birth month were evaluated to predict lemur longevity.
-The dataset was divided into wild-born and captive-born lemurs to understand the differences in lifespan between the two environments and separate models were built for each group.
+The estimand for our model is the age or the lifespan of the lemur. The predictors are `Sex`, `Species`, `Genus`, and `month_born`.
+
+Using Generalized Linear Models (GLMs) with Gaussian distribution, factors including sex, species, genus, and birth month were evaluated to predict lemur longevity in wild and captive.
+To do this, the dataset was divided into wild-born and captive-born lemurs to understand the differences in lifespan between the two environments and separate models were built for each group.
 
 The findings reveal that lemurs in captivity exhibit significantly longer lifespans.
-In wild populations, males tend to live longer than females, while species differences are prominent in both environments, emphasizing the need for species-specific conservation strategies.
-Birth month was found to have inconsistent effects across both groups, suggesting complex interactions with environmental conditions.
+Males tend to live longer than females in both captive and wild environments, however the difference is more pronounced in captivity.
+Species-specific variations in lifespan were observed, with some species thriving in captivity while others perform better in the wild.
+Birth month effects were inconsistent, suggesting complex interactions with environmental factors.
 
@@
 Finally, @sec-appendix is the appendix and includes additional information on data cleaning, the analysis dataset, and the model dataset.
+@sec-appendix-surveys includes the details of surveys and observational data.
 
@@
 These packages include `tidyverse` [@cite-tidyverse], `ggplot2` [@cite-ggplot], `dplyr` [@cite-dplyr], `patchwork` [@cite-patchwork], `arrow` [@cite-arrow], `knitr` [@cite-knitr], `rstanarm` [@cite-rstanarm] and `kableExtra` [@cite-kableExtra].
+`testthat` [@cite-testthat] was used for testing the simulated and actual datasets.
 
@@
 Finally, we have the target variable `Age` which is the lifespan of the lemur in years. It is a continuous variable.
+
+@cookson_duke_lemur_center also had another dataset which contained the weight of these lemurs, but I did not use it in this analysis.
+It includes details into pregnancy and weights of lemur during ther lifetime, it can be added to see the how trends in weight, pregnancy and litter size can affect the lifespan of lemurs.
 
@@
 
-wild_plot_species <- create_plot(wild_data, "species", WILD_COLOR, "Wild Lemurs", "Species")
-captive_plot_species <- create_plot(captive_data, "species", CAPTIVE_COLOR, "Captive Lemurs", "Species")
+wild_plot_species <- create_plot(wild_data, "species", WILD_COLOR, "", "Species")
+captive_plot_species <- create_plot(captive_data, "species", CAPTIVE_COLOR, "", "Species")
 
-wild_plot_genus <- create_plot(wild_data, "genus", WILD_COLOR, "Wild Lemurs", "Genus")
-captive_plot_genus <- create_plot(captive_data, "genus", CAPTIVE_COLOR, "Captive Lemurs", "Genus")
+wild_plot_genus <- create_plot(wild_data, "genus", WILD_COLOR, "", "Genus")
+captive_plot_genus <- create_plot(captive_data, "genus", CAPTIVE_COLOR, "", "Genus")
 
-wild_plot_month <- create_plot(wild_data, "month_born", WILD_COLOR, "Wild Lemurs", "Birth Month")
-captive_plot_month <- create_plot(captive_data, "month_born", CAPTIVE_COLOR, "Captive Lemurs", "Birth Month")
+wild_plot_month <- create_plot(wild_data, "month_born", WILD_COLOR, "", "Birth Month")
+captive_plot_month <- create_plot(captive_data, "month_born", CAPTIVE_COLOR, "", "Birth Month")
 ```
@@
 #| label: fig-sex-plots
-#| fig-cap: Sex distribution of wild and captive lemurs represented by bar plots for female (F) and male (M) individuals.
+#| fig-cap: Sex distribution of wild and captive lemurs for female (F) and male (M) individuals. 
 # Display plots
@@
 #| label: fig-month-plots
-#| fig-cap: Distribution of lemurs by the month they were born in. On left is the distribution for lemurs in the wild and on the right is for lemurs in captivity.
+#| fig-cap: Distribution of lemurs by the month they were born in the wild (on left) and in captivity (on right).
 wild_plot_month + captive_plot_month
@@
 #| label: fig-age-distribution
-#| fig-cap: Distribution of ages of lemurs in the dataset for lemurs in captivity (on right) and in the wild (on left).
+#| fig-cap: Distribution of ages of lemurs in the dataset in wild (on left) and in captivity (on right).
 
@@
 
-@sec-appendix-model covers coefficients, standard errors, convergence checks and other model diagnostics of our model.
+@sec-appendix-model covers coefficients and standard errors and @#sec-appendix-ppc includes posterior predictive checks for the model.
 
@@
 
+# Survey, Sampling and Observational Data {#sec-appendix-surveys}
+
+The study relies on observational data collected by the Duke Lemur Center (DLC), which documents lemur lifespans in wild and captive settings.
+Unlike experimental data, observational datasets like this are not generated through controlled interventions.
+This limits the ability to infer causal relationships.
+
+## Data Collection Methodology
+
+The Duke Lemur Center employed standard operating procedures to gather data, adhering to USDA, AZA, and IACUC guidelines.
+The original data was handwritten and later digitized into two primary databases: the Animal Record Keeping System (ARKS) and MedARKS, and subsequently the Zoological Information Management System (ZIMS).
+These steps ensured consistency and traceability but may have introduced biases, particularly during digitization.
+More on this can be found in @sec-data_collection.
+
+## Sampling Frame
+
+The dataset includes both wild and captive lemurs, but the sampling frame is imbalanced, with far fewer wild lemurs.
+This is because it is difficult to track wild populations compared to the captive ones.
+This can be conutered by increasing surveillance and tracking of wild populations by installing cameras and marking them for easy identification.
+Alternatively, some study can be conducted to distinguish different lemurs through some of their unique biological featrues.
+For example, tigers are identified by their stripes and pug marks.
+
+To address this, statistical adjustments such as weighting could be applied, though this study opted for separate models to analyze wild and captive populations independently.
+
+## Sources of Bias
+
+1. Survivor Bias: Only lemurs with complete lifecycle information are included, potentially skewing lifespan estimates toward older individuals.
+
+2. Selection Bias: The focus on lemurs cared for by the DLC may not represent broader populations, especially in the wild.
+
+3. Measurement Error: Data inconsistencies could arise from manual transcription and digitization of historical records.
+
+## Literature Context
+
+The methodologies align with best practices in ecological studies, as outlined by Zehr et al.
+(2014), who emphasize the importance of longitudinal tracking and database integration for species conservation.
+The use of generalized linear models (GLMs) aligns with approaches recommended in similar studies to account for categorical and continuous predictors.
+
+## Key Takeaways
+
+The study’s observational nature limits causal inference but provides valuable insights into factors influencing lemur longevity.
+Careful data cleaning and validation steps mitigate some biases, but limitations persist, particularly regarding the underrepresentation of wild populations.
+Future work could employ advanced sampling techniques or collaborative data pooling with other conservation organizations to improve generalizability.
+
 # Data Cleaning {#sec-appendix-data-cleaning}

### 2024-12-04 scripts/00-install_packages.r

@@ -13,7 +13,10 @@ packages <- c(
   "knitr",
   "ggplot2",
   "arrow",
-  "rstanarm"
+  "rstanarm",
+  "testthat",
+  "kableExtra",
+  "patchwork"
 )
 
 # Install packages if they are not already installed

### 2024-12-04 scripts/01-simulate_data.r

@@ -4,7 +4,7 @@
 # Date: 18 November 2024
 # Contact: khushaal.nandwani@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: None
+# Pre-requisites: Rnu 00-install-packages.r
 
 ## Workspace Setup ##
 library(tidyverse)

### 2024-12-04 scripts/02-download_data.r

@@ -1,3 +1,11 @@
+#### Preamble ####
+# Purpose: Downloads the data file from the internet
+# Author: Khushaal Nandwani
+# Date: 18 November 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+
 url <- "https://raw.githubusercontent.com/tacookson/data/master/duke-lemur-center/animals.txt"
 
 destination_dir <- "data/raw_data"

### 2024-12-04 scripts/03-clean_data.r

@@ -4,7 +4,8 @@
 # Date: 18 November 2024
 # Contact: khushaal.nandwani@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The raw data file should be present in the data/raw_data folde
+# Pre-requisites: The raw data file should be present in the data/raw_data folder. 
+#                 If not, run 02-download_data.r to download it.
 
 ## Workspace Setup ##
 library(tidyverse)

### 2024-12-04 scripts/04-test-data.r

@@ -1,112 +1,0 @@
-#### Preamble ####
-# Purpose: Tests the analysis data
-# Author: Khushaal Nandwani
-# Date: 27 November 2024
-# Contact: khushaal.nandwani@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: Run 03-clean_data.r
-
-## Workspace setup ##
-library(tidyverse)
-library(arrow)
-
-## Tests ##
-print("Testing analysis data for wild and captive lemurs")
-
-wild_data <- read_parquet("data/analysis_data/wild.parquet")
-captive_data <- read_parquet("data/analysis_data/captive.parquet")
-
-all_tests_passed = TRUE
-
-# Test 1: Check for any NA values or missing data
-if (any(is.na(wild_data))) {
-  print("Test 1 failed: Missing values in wild data")
-  all_tests_passed = FALSE
-}
-
-if (any(is.na(captive_data))) {
-  print("Test 1 failed: Missing values in captive data")
-  all_tests_passed = FALSE
-}
-
-# Test 2: Check number of columns
-if (ncol(wild_data) != 7) {
-  print("Test 2 failed: Incorrect number of columns in wild data")
-  all_tests_passed = FALSE
-}
-
-if (ncol(captive_data) != 7) {
-  print("Test 2 failed: Incorrect number of columns in captive data")
-  all_tests_passed = FALSE
-}
-
-# Test 3: Check age is between 0 and 50
-if (any(wild_data$age < 0) | any(wild_data$age > 50)) {
-  print("Test 3 failed: Age values out of range in wild data")
-  all_tests_passed = FALSE
-}
-
-# Test 4: Check sex is either M or F
-if (any(!(wild_data$sex %in% c("M", "F")))) {
-  print("Test 4 failed: Invalid sex in wild data")
-    all_tests_passed = FALSE
-}
-
-if (any(!(captive_data$sex %in% c("M", "F")))) {
-  print("Test 4 failed: Invalid sex in captive data")
-    all_tests_passed = FALSE
-}
-
-# Test 5: Check birth type is either captive or wild
-if (any(!(wild_data$birth_type %in% c("captive", "wild")))) {
-  print("Test 5 failed: Invalid birth type in wild data")
-  all_tests_passed = FALSE
-}
-
-if (any(!(captive_data$birth_type %in% c("captive", "wild")))) {
[…]

### 2024-12-04 scripts/04a-test-sim-data.r

@@ -1,0 +1,109 @@
+#### Preamble ####
+# Purpose: Tests the simulated data
+# Author: Khushaal Nandwani
+# Date: 18 November 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: Run 01-simulate_data.r
+
+
+## Workspace Setup ## 
+library(testthat)
+library(tidyverse)
+
+# Load the dataset
+data <- read_csv("data/analysis_data/sim_data.csv")
+
+# Start writing tests
+test_that("Dataset structure is correct", {
+  # Test the column names
+  expect_named(data, c(
+    "animal_id", "species", "sex", "birth_type", 
+    "litter_size", "mother_father_same", "age", "month_born"
+  ))
+
+  # Test the number of rows
+  expect_equal(nrow(data), 1200)
+
+  # Test the number of columns
+  expect_equal(ncol(data), 8)
+})
+
+test_that("Animal IDs are unique and sequential", {
+  expect_true(all(data$animal_id == 1:1200))
+})
+
+test_that("Species values are valid", {
+  types_of_species <- c(
+    "Northern greater galago",
+    "Collared brown lemur",
+    "Eulemur hybrid",
+    "Red-fronted brown lemur",
+    "Mohol bushbaby",
+    "Eastern lesser bamboo lemur",
+    "Black lemur",
+    "Common brown lemur"
+  )
+  expect_true(all(data$species %in% types_of_species))
+})
+
+test_that("Sex values are valid", {
+  expect_true(all(data$sex %in% c("M", "F")))
+})
+
+test_that("Birth type values are valid", {
+  expect_true(all(data$birth_type %in% c("captive", "wild")))
+})
+
+test_that("Litter size is within the expected range", {
+  expect_true(all(data$litter_size >= 1 & data$litter_size <= 5))
+})
+
+test_that("Mother-father same values are binary", {
+  expect_true(all(data$mother_father_same %in% c(0, 1)))
+})
+
+test_that("Age values are reasonable based on birth type", {
+  captive_ages <- data$age[data$birth_type == "captive"]
+  wild_ages <- data$age[data$birth_type == "wild"]
+  
+  # Check captive ages
+  expect_true(all(captive_ages >= 0))
+  expect_true(mean(captive_ages) > mean(wild_ages)) # Captive ages should generally be higher
+
+  # Check wild ages
[…]

### 2024-12-04 scripts/04b-test-actual-data.r

@@ -1,0 +1,82 @@
+#### Preamble ####
+# Purpose: Tests the analysis data
+# Author: Khushaal Nandwani
+# Date: 27 November 2024
+# Contact: khushaal.nandwani@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: Run 03-clean_data.r
+
+## Workspace setup ##
+library(testthat)
+library(tidyverse)
+library(arrow)
+
+# Load data
+wild_data <- read_parquet("data/analysis_data/wild.parquet")
+captive_data <- read_parquet("data/analysis_data/captive.parquet")
+
+# Define valid values for checks
+valid_genus <- c("O", "E", "H", "G", "V", "L", "P", "C", "M", "N", "D")
+valid_species <- c("GG", "COL", "MOH", "MON", "VV", "UL", "RUB", "CAT", "RUF", "MAC", "FUL", "AR", "COQ", "MED", "MUR", "COU", "TAR", "PYG", "ZAZ", "ALB", "COR", "SAN", "FLA", "MAD", "POT")
+valid_months <- month.abb
+
+# Tests
+test_that("wild data has no missing values", {
+  expect_false(any(is.na(wild_data)), info = "Missing values in wild data")
+})
+
+test_that("captive data has no missing values", {
+  expect_false(any(is.na(captive_data)), info = "Missing values in captive data")
+})
+
+test_that("wild data has correct number of columns", {
+  expect_equal(ncol(wild_data), 7, info = "Incorrect number of columns in wild data")
+})
+
+test_that("captive data has correct number of columns", {
+  expect_equal(ncol(captive_data), 7, info = "Incorrect number of columns in captive data")
+})
+
+test_that("wild data age is between 0 and 50", {
+  expect_true(all(wild_data$age >= 0 & wild_data$age <= 50), info = "Age values out of range in wild data")
+})
+
+test_that("wild data sex is valid", {
+  expect_true(all(wild_data$sex %in% c("M", "F")), info = "Invalid sex in wild data")
+})
+
+test_that("captive data sex is valid", {
+  expect_true(all(captive_data$sex %in% c("M", "F")), info = "Invalid sex in captive data")
+})
+
+test_that("wild data birth_type is valid", {
+  expect_true(all(wild_data$birth_type %in% c("captive", "wild")), info = "Invalid birth type in wild data")
+})
+
+test_that("captive data birth_type is valid", {
+  expect_true(all(captive_data$birth_type %in% c("captive", "wild")), info = "Invalid birth type in captive data")
+})
+
[…]

### 2024-12-04 scripts/05-data-analysis.r

@@ -5,7 +5,7 @@
 # Date: 18 November 2024
 # Contact: khushaal.nandwani@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: None
+# Pre-requisites: Run 04b-test-actual-data.r
 
 ## Workspace setup ##
 library(arrow)

### 2024-12-04 scripts/06-predictions.r

@@ -4,7 +4,7 @@
 # Date: 18 April 2024
 # Contact: khushaal.nandwani@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Run 04-data-analysis.r
+# Pre-requisites: Run 05-data-analysis.r
 
 ## Workspace setup ##
 library(tidyverse)

### 2024-12-04 scripts/06-predictions.r
predictions script updated

@@ -11,7 +11,10 @@ library(tidyverse)
 library(arrow)
 
 # Prediction datasets
-analysis_data <- read_csv("data/analysis_data/wild.csv")
+# We are only using wild data for predictions as certain 
+# species are only present in the captive, which could disrupt the wild model.
+analysis_data <- read_parquet("data/analysis_data/wild.parquet")
+
 
 # Load models
 wild_model <- readRDS("models/wild_model.rds")
@@ -46,4 +49,3 @@ predictions_df <- data.frame(
 
 # save the predictions
 write_csv(predictions_df, "data/predictions/predictions.csv")
-

### 2024-12-15 paper/paper.qmd
prose refinedand frontend mentioned in paper

@@
 abstract: "This study investigates the lifespan of lemurs in wild and captive environments using data from the Duke Lemur Center, aiming to identify how factors such as sex, species, genus, and birth month influence longevity.
-Using Generalized Linear Models, the analysis highlights that lemurs in captivity live significantly longer than their wild counterparts, emphasizing the impact of conservation practices.
-The findings show species-specific variations in lifespan and the pronounced influence of sex differences in captivity.
-These insights help identify effective practices in captivity and suggest areas for improvement to optimize conservation strategies, contributing to the survival of this endangered mammalian group."
+Using Generalized Linear Models, the analysis found that lemurs in captivity live significantly longer than their wild counterparts, emphasizing the impact of conservation practices.
+The study also identified species-specific variations in lifespan and a pronounced influence of sex in captivity.
+The findings from this study can help determine what we are doing right and where we are going wrong."
 format: pdf
@@
 Lemurs are the world’s most endangered group of mammals (@citedukelemurcenter).
-Understanding the factors that influence their lifespan is crucial for guiding effective conservation efforts.
+Understanding the factors that influence their lifespan is important for guiding effective conservation efforts.
 The Duke Lemur Center (DLC), established in 1966, has played a pivotal role in the study and preservation of lemurs, maintaining the largest population of these species outside their native region Madagascar.
@@
 My study uses a dataset from the DLC to analyze if and how sex, species, genus and month a lemur is born in can affect its lifespan in the wild and captivity.
-This study can provide insights into which lemur species we are struggling to conserve in captivity and their performance in the wild.
+This study can help us understand which lemur species we are struggling to conserve in captivity and their corresponding performance in the wild.
 It will highlight areas where our conservation efforts have been effective.
@@
 
-The findings reveal that lemurs in captivity exhibit significantly longer lifespans.
+The findings show that lemurs in captivity exhibit significantly longer lifespans.
 Males tend to live longer than females in both captive and wild environments, however the difference is more pronounced in captivity.
@@
 
-### Data Acquisition
+### Data Acquisition {#sec-data-acquisition}
  As @3zehr points out, DLC staff collected data about the lemurs according to standard operating procedures and USDA, AZA, and IACUC guidelines.
@@
 
-### Data Processing
+### Data Processing {#sec-data-processing}
 The DLC used SAS software to build a database for the lemur data.
@@
 At the expense of, not being able to study the interaction effects of `birth_type` with other predictors.
+
+## Model API and Frontend
+The model API code along with its instructions is available on the GitHub repository [https://github.com/khushaal-nandwani/predicting-lemurs-api](https://github.com/khushaal-nandwani/predicting-lemurs-api).
+Please read the Readme file in the repository for more information on how to call the API.
+
+There is also a frontend developed for the api, where one can dynamically enter the `sex`, `genus`, `species` and `month_born` of the lemur and can get predictions for that lemur's lifespan in both wild and captive enviorments.
+The frontend is live at [http://3.99.183.165:5173/](http://3.99.183.165:5173/).
+The code for the frontend is available at [https://github.com/khushaal-nandwani/predicting-lemurs-frontend](https://github.com/khushaal-nandwani/predicting-lemurs-frontend).
+
+The API is built using FastAPI and deployed on an AWS EC2 Instance. The frontend app is made in Vue and deployed on the same EC2 Instance.
 
@@
 For months in @fig-month-born-effects, we see captive lemurs outlive wild lemurs most of the months except August, January, March, Novemeber and September.
-It is worth exploring the reasons behind these differences by Duke Lemur Center, as they can provide insights into the impact of seasonal factors on lemur lifespan.
+It is worth exploring the reasons behind these differences by Duke Lemur Center, as they can help us understand the impact of seasonal factors on lemur lifespan.
 
@@
 ## Key Findings 
-One key insight is that captivity significantly increases lemur lifespan.
+We learn that captivity significantly increases lemur lifespan.
 Controlled environments provide consistent access to food, medical care, and protection from predators, reducing environmental stresses that contribute to shorter lifespans in the wild.
@@
 
-Species-specific differences in lifespan emerge as a critical factor.
+Species-specific differences in lifespan emerge as a factor.
 Most of the species thrive in captivity due to tailored conservation practices, while some including `ALB`, `COR`, `MAC`, and `MON` perform better in the wild.
@@
 ## Drawbacks of the Study
-While the study provides valuable insights, it has its limitations. First, the dataset is skewed towards captive lemurs, limiting the robustness of wild lifespan analysis.
+Despite some useful findings, the study has its limitations. First, the dataset is skewed towards captive lemurs, limiting the robustness of wild lifespan analysis.
 
-Second, the exclusion of interactions between birth type and other predictors in the combined model reduces the depth of insights.
+Second, the exclusion of interactions between birth type and other predictors was a tradeoff worth reconsidering.
 
@@
 
-6. Using advanced modeling techniques like Bayesian models or neural networks for more nuanced predictions.
+6. Using modeling techniques like Bayesian models or neural networks to improve predictive accuracy.
 
@@
 These steps ensured consistency and traceability but may have introduced biases, particularly during digitization.
-More on this can be found in @sec-data_collection.
+More on how the data was acquired can be found in @sec-data-acquisition.
 
@@
 The use of generalized linear models (GLMs) aligns with approaches recommended in similar studies to account for categorical and continuous predictors.
+More on how the data was processed can be found in @sec-data-processing.
 
@@
 
-The study’s observational nature limits causal inference but provides valuable insights into factors influencing lemur longevity.
+The study’s observational nature limits causal inference but helps us understand factors influencing lemur longevity.
 Careful data cleaning and validation steps mitigate some biases, but limitations persist, particularly regarding the underrepresentation of wild populations.
-Future work could employ advanced sampling techniques or collaborative data pooling with other conservation organizations to improve generalizability.
+Future work could employ other sampling techniques or collaborative data pooling with other conservation organizations to improve generalizability.
 

### 2024-12-15 paper/paper.qmd
abstract improved

@@
 The study also identified species-specific variations in lifespan and a pronounced influence of sex in captivity.
-The findings from this study can help determine what we are doing right and where we are going wrong."
+The findings from this study can help determine what we are doing right and where we are going wrong in our conservation efforts for this endangered species."
 format: pdf

### 2024-12-15 paper/paper.qmd
Minor detail added in estimand paragraph
ref.bib renamed to references.bib

--- 2024-12-15 paper/paper.qmd
@@
 toc: true
-bibliography: ref.bib
+bibliography: references.bib
 # editor:
@@
 
-The estimand for our model is the age or the lifespan of the lemur. The predictors are `Sex`, `Species`, `Genus`, and `month_born`.
+The estimand for our model is the age or the lifespan of the lemur in wild and captive enviornments. The predictors are `Sex`, `Species`, `Genus`, and `month_born`.
 
